---
title: "Neural likelihoods"
description: "Using and validating neural-network likelihoods in EMC2"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{"Neural likelihoods"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(EMC2)
```

## Loading a network is not the same as being able to infer with it

*EMC2* can use a trained neural network in place of a model's likelihood:
the shipped normalizing flows `DDMnn` (the full DDM) and `RDMnn` (the racing
diffusion model), and any network registered with `register_nn_model()`
(flows, networks that regress the log-likelihood, and likelihood
approximation networks converted from ONNX). Such a model is used exactly like
`DDM` or `RDM`: it goes into `design()`, gets a prior, is fitted.

A network is an approximation, trained on simulations from part of the
parameter space. Its error is not a constant: where it varies with the
parameters it moves the posterior, and a model that looks accurate point by
point can still be biased where it matters for inference. So a network has to
be checked, against the exact likelihood, before its posteriors are believed.
This vignette describes the checks *EMC2* provides and what they found for the
shipped networks. The reasoning behind them, and the mistakes they guard
against, are documented in the NLE project's handover notes (Wuth, 2026:
`METHODS.md` section 6, "What the diagnostics can and cannot do", and
`TRAPS.md`); they are not restated here.

## A cell: the network and its control, differing in one respect

Every check runs on a *cell*: one design with independent normal priors on
the sampled scale, built twice, once with the network and once with an
analytic *control*, by default the model the network approximates (for
`DDMnn`, `DDM`). The two cells share the design, constants, defaults, prior
and simulator; they differ in the likelihood only. The same check on the
control must come out at about zero; what it does show is not the network's
doing.

```{r}
cell <- nn_cell(DDMnn,
  mean = c(v = 1, a = log(.8), t0 = log(.3), Z = qnorm(.5), SZ = qnorm(.2),
           sv = log(.1), st0 = log(.05)),
  sd = c(v = .15, a = .15, t0 = .1, Z = .05, SZ = .15, sv = .15, st0 = .15))
cell
```

Priors are always named (never matched by position). Parameters not given a
prior stay at the model's default, which must lie inside the network's
training region (`DDMnn` refuses `sv = SZ = st0 = 0`, `DDM`'s defaults, because
the network was not trained there). The within-trial noise `s` is held at its
default by default, as it sets the scale.

**The prior must stay inside the training region.** The network rejects
parameter vectors outside the region it was trained on, which truncates the
posterior there. For an SBC that is fatal: a replicate whose true parameters
lie outside cannot be recovered. `nn_cell()` refuses a prior whose mean +- 4 sd
leaves the region, and `run_sbc()` makes the same check (`nn_in_box()`) for
every neural-likelihood design:

```{r, error = TRUE}
des <- design(factors = list(subjects = 1), Rlevels = c("a", "b"), model = RDMnn,
              formula = list(v ~ 0 + lR, B ~ 1, t0 ~ 1, A ~ 1), constants = c(s = log(1)),
              report_p_vector = FALSE)
pri <- prior(des, type = "single",
             pmean = c(v_lRa = log(2), v_lRb = log(1), B = 0, t0 = log(.2), A = log(.3)),
             psd = c(v_lRa = .1, v_lRb = .1, B = .1, t0 = .5, A = .1))
nn_in_box(des, pri)
```

## Four checks, cheapest first

```{r, eval = FALSE}
nn_total_mass(cell)                        # seconds
nn_score_bias(cell, n_draws = 200)         # seconds
nn_posterior_shift(cell, n_datasets = 60)  # about a minute
nn_sbc_cell(cell, trials = 400, replicates = 500, cores_per_chain = 10,
            archive_dir = "2026-10-01_dev-nle_cap256w-sbc")   # hours
```

**Total mass** (`nn_total_mass()`) integrates the network's likelihood over rt
and responses for parameter vectors drawn from the prior. A flow integrates to
one by construction; a network that regresses the log-likelihood does not,
which is harmless while the log mass is constant in the parameters (it cancels
from the posterior). The same integral of the control on the same grid
separates the network from the grid. The column `before_support` is the mass
the network puts before the non-decision time, where the exact density is
zero. A flow's support is all rt > 0, so it smears the leading edge of the
distribution a little, the region that informs the non-decision time.

**Score bias** (`nn_score_bias()`) is the expected score of the network's
log-likelihood at the true parameters, per trial: zero for an exact
likelihood, and the direction in which the network pulls the posterior
otherwise. It is computed by quadrature over rt (weighted by the control's
likelihood) and a central difference, so the control's own score bias must be
about zero; if it is not, the estimator is at fault, not the network.
Parameters that move the support (`t0`) are excluded. Score bias is local: it
ranks networks of one architecture but does not predict SBC bias across
architectures. It is necessary, not sufficient.

**Posterior shift** (`nn_posterior_shift()`) computes what an SBC cell
computes, without the sampler: for data sets simulated from the prior, the
posterior under the network and under the control (Laplace approximations
refined by importance sampling), and from them the standardised bias of each
posterior mean, `mean(post mean - truth) / sd(post mean - truth)`, the shift
of the network's posterior from the control's in posterior sds, and the ratio
of posterior sds. The control column must be within Monte Carlo error of zero
(standard error about `1 / sqrt(n)`); the paired shift is much more precise
than either column. In the NLE project it predicted the standardised bias of
500-replicate SBC cells within about 0.1. Use it to screen, not to certify.

**An SBC cell with its control** (`nn_sbc_cell()`) is the confirmation: it
runs `run_sbc()` on the network's cell and on the control's, summarises both
(`nn_sbc_summary()`: standardised bias, with the gate |value| <= 0.5; coverage;
the KS test of the ranks; the ECDF envelope), and, given `archive_dir`, writes
the run in the layout of the EMC2 SBC archive. With `run = FALSE` it only
writes the bundle (the cell and a script that runs it), to be copied to a
cluster.

None of these checks identifiability: SBC passes by construction for a
parameter the data do not inform. For a new model, add parameter recovery
with the prior centred away from the truth, and likelihood profiles.

## What the checks found for the shipped networks

The cells are those the NLE project used (the prior above for the DDM
networks; for `rdm_small`, a two-accumulator race with `v ~ lM` and prior
means `v = log(1.5)`, `v_lMTRUE = 0.4`, `B = 0`, `A = log(0.3)`,
`t0 = log(0.2)`, sds 0.1). The script
`system.file("scripts", "nle_validation_kit.R", package = "EMC2")`
reproduces the numbers.

All numbers are on the sampled scale (log for `a`, `t0`, `sv`, `st0`, `B`,
`A`, the rates of the race; probit for `Z`, `SZ`). "NLE" is the value the
project recorded with its own scripts, from other draws (except where noted).

**Total mass** (50 prior draws). The three flows integrate to one within the
quadrature error: mean `log Z` below 2e-5 in absolute value, sd below 1.1e-5,
and they put at most 1.6e-5 of their mass before the non-decision time. The
LAN (a network regressing the log-likelihood, converted from ONNX) is not
normalised by construction: sd of `log Z` 0.003 over the prior, similar to the
project's own regression network (0.002). Before its support it returns its
training floor, log density -16.1, where the exact density is zero (a fit
floors it at log(1e-10) = -23).

**Score bias** (h = 0.01; mean over 200 prior draws, standard error in
brackets; LAN over 80 draws, rt grid to 60 s). Every control is below 5e-4 in
absolute value.

| network | parameter | g (se) | NLE |
|---|---|---|---|
| `ddm_cap256w_c4` | v | +0.0001 (0.0001) | +0.0004 |
| | a | +0.0153 (0.0014) | +0.0118 |
| | Z | -0.0061 (0.0005) | -0.0063 |
| | SZ | -0.0017 (0.0001) | -0.0017 |
| | sv | -0.0001 (0.0000) | -0.0001 |
| | st0 | +0.0038 (0.0004) | +0.0039 |
| `ddm_st0zero` | v | -0.0017 (0.0001) | -0.0016 |
| | a | +0.0338 (0.0010) | +0.0336 |
| | Z | +0.0008 (0.0004) | +0.0014 |
| | SZ | -0.0017 (0.0001) | -0.0017 |
| `rdm_small` | v | -0.0039 (0.0003) | -0.0035 |
| | v_lMTRUE | -0.0021 (0.0002) | -0.0018 |
| | B | +0.0038 (0.0004) | +0.0030 |
| | A | +0.0009 (0.0001) | +0.0008 |
| LAN `ddm_uniform_st` | v | +0.0000 (0.0024) | +0.0038 |
| | a | -0.1098 (0.0037) | -0.1221 |
| | z | -0.0094 (0.0027) | +0.0025 |

The race control reproduces the project's (-0.00012 on v, -0.00021 on B) to
every printed digit. For the LAN the grid has to reach 60 s: with the default
8 s the control's g on `a` is -0.005, because the region includes slow
parameter vectors with probability beyond 8 s (the print reports how much of
the probability the grid holds).

**Posterior shift** (60 data sets x 400 trials, 2000 draws; about a minute for
all four networks on 10 cores). Network standardised bias, with the shift
from the control's posterior in control posterior sds in brackets; every
control column is within +-0.36 (standard error 0.13).

| network | parameter | network (shift) | SBC, 500 replicates (NLE) |
|---|---|---|---|
| `ddm_cap256w_c4` | v | +0.03 (+0.04) | +0.05 |
| | a | -0.05 (0.00) | +0.01 |
| | t0 | +0.25 (+0.15) | +0.32 |
| | Z | +0.08 (-0.05) | -0.03 |
| | SZ | +0.03 (+0.01) | -0.05 |
| | sv | -0.05 (+0.01) | +0.01 |
| | st0 | -0.22 (0.00) | -0.14 |
| `ddm_st0zero` | v | +0.03 (-0.07) | |
| | a | +0.31 (+0.16) | |
| | t0 | +0.11 (+0.20) | +0.28 |
| | Z | +0.11 (+0.11) | worst, +0.28 |
| | SZ | +0.06 (+0.04) | |
| | sv | -0.05 (+0.01) | |
| `rdm_small` | v | +0.11 (+0.03) | +0.03 |
| | v_lMTRUE | -0.20 (-0.01) | -0.03 |
| | B | -0.05 (+0.12) | +0.04 |
| | A | +0.37 (+0.02) | +0.09 |
| | t0 | -0.09 (-0.16) | 0.00 |
| LAN `ddm_uniform_st` | v | +0.09 (+0.08) | |
| | a | **-0.81 (-0.78)** | |
| | z | -0.31 (-0.17) | |
| | t | +0.08 (+0.12) | |
| | st | 0.00 (-0.04) | |

For `ddm_st0zero` the project ran the same check with the same seeds and the
same (exact) simulator: every value agrees within 0.004. The race network's
larger network values (v_lMTRUE, A) are shared by its control (-0.20, +0.36):
they come from the 60 true parameter vectors, not the network, which is why
the paired shift is the column to read.

In short: the DDM flows are calibrated within the SBC gate (|standardised
bias| <= 0.5) at 400 trials, with a positive non-decision-time bias as their
main error; the race flow is calibrated; the LAN underestimates boundary
separation by about 0.8 posterior sds at 400 trials and should not be used
for inference in this region without further work.

## Practical notes

- The checks and `run_sbc()` run their work in forked processes (`cores`,
  `cores_per_chain`). On macOS, *EMC2* keeps Apple's Accelerate library (the
  BLAS its compiled code uses there) single-threaded by setting
  `VECLIB_MAXIMUM_THREADS=1` when it is loaded, because its multithreaded
  matrix products crash forked processes. On Linux with OpenBLAS, set
  `OPENBLAS_NUM_THREADS=1` for the same reason (and to avoid
  oversubscription).
- A network registered from a file path must be at the same path wherever
  the cell is run; the shipped networks are found by name.
