---
title: "Rating Models"
description: "Choice, response time and confidence ratings with the correlated multiple-threshold LNR"
author: "Andrew Heathcote"
output: rmarkdown::html_document
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{"Rating Models"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
rm(list = ls())
library(EMC2)
set.seed(30)
```

## Introduction

Rating tasks add a discrete confidence (or other ordered) rating to a choice
response time task: on every trial the data contain not just a choice `R` and
a response time `rt`, but also a rating `RR`, coded as an integer from 1
(the lowest rating) to `K` (the highest).

This vignette introduces the first rating model in *EMC2*, the correlated
multiple-threshold log-normal race (MTLNR) of @reynolds2020correlated. The
MTLNR extends the log-normal race (`?LNR`) to two *correlated* accumulators
and reads the confidence rating off the losing accumulator: at the moment the
winner finishes, the loser's state (as a proportion of its own threshold) is
compared against `K - 1` further thresholds to give one of `K` ratings. This
is a natural model for tasks such as old/new recognition memory with a
confidence judgment, the running example below.

The steps are:

1. Specify a rating-task design and inspect the threshold mapping
2. Simulate data and summarize it by choice and rating
3. Define priors and fit the model
4. Check the fit against the truth and the data (posterior predictives)
5. A rating-specific diagnostic: the z-transformed ROC

For the full parameterization see `?MTLNR`.

## 1. Specify a Rating-Task Design

We simulate a two-choice recognition-memory task: participants see a word
that was either `new` or `old`, and respond `new` or `old` with a rating of
their confidence on a 3-point scale (`RR` = 1 low, 3 high). `MTLNR()` is a
factory: `MTLNR(n_ratings = 3)` (the default) builds the model for 3 ratings.

As with the other race models (see `vignette("race-models")`), `matchfun`
flags whether the accumulator's response label matches the true word type,
giving the latent match factor `lM`; here it lets accuracy shift the mean log
finishing time. The two confidence criteria `c1` and `c2` are mapped `~ lR`
(one pair per response) since confidence thresholds belong to the *losing*
accumulator and so can differ between "new" and "old" responses, capturing a
response-specific bias in how confidence is reported.

```{r}
design_mtlnr <- design(
  factors = list(subjects = 1, S = c("new", "old")),
  Rlevels = c("new", "old"),
  matchfun = function(d) d$S == d$lR,
  formula = list(m ~ lM, s ~ 1, t0 ~ 1, rho ~ 1, c1 ~ lR, c2 ~ lR),
  model = MTLNR
)
sampled_pars(design_mtlnr)
```

## 2. Rating Thresholds and Their Mapping

The criteria `c1`, `c2` are sampled on a log scale that guarantees the
underlying thresholds are ordered; `mapped_pars()` reports the natural-scale
thresholds `d1 < d2` (as proportions of the losing accumulator's own
threshold) alongside `rho`, mapped by response:

```{r}
p_vector <- sampled_pars(design_mtlnr)
p_vector[] <- c(m = -.5, m_lMTRUE = -.6, s = log(.8), t0 = log(.3), rho = qnorm(.75),
                c1 = log(.4), c1_lRold = .1, c2 = log(.6), c2_lRold = -.2)
mapped_pars(design_mtlnr, p_vector)
```

`rho` (here mapped through a scaled probit to lie in (-1, 1), `?MTLNR`) is
shared by both accumulators and is `r round(2 * pnorm(p_vector[["rho"]]) - 1, 2)`
for these parameter values: a strong positive trial-to-trial correlation
between the two finishing times.

## 3. Simulate Data and Summarize by Choice and Rating

```{r, results = "hide"}
dat <- make_data(parameters = p_vector, design = design_mtlnr, n_trials = 1000)
head(dat)
```

`rating_summary()` gives, for each folded response (choice crossed with
rating), the response proportion and response-time quantiles:

```{r}
rating_summary(dat, factors = "S")
```

`plot_ratings()` plots the same information, response proportions on top and
RT quantiles below, split by word type:

```{r, fig.alt = "Response proportions and RT quantiles by confidence rating for new and old words"}
plot_ratings(dat, factors = "S")
```

## 4. Set Priors and Fit

```{r, results = "hide"}
prior_mtlnr <- prior(
  design = design_mtlnr, type = "single",
  pmean = c(m = 0, m_lMTRUE = 0, s = log(.7), t0 = log(.25), rho = 0,
            c1 = log(.5), c1_lRold = 0, c2 = log(.5), c2_lRold = 0),
  psd = c(m = 1, m_lMTRUE = 1, s = .5, t0 = .4, rho = 1,
          c1 = .5, c1_lRold = .5, c2 = .5, c2_lRold = .5)
)
```

```{r, fig.alt = "Prior densities for the recognition-memory MTLNR example"}
plot(prior_mtlnr, N = 1e3)
```

```{r, results = "hide"}
emc <- make_emc(dat, design_mtlnr, prior_list = prior_mtlnr, type = "single")
```

```{r, eval = FALSE}
emc <- fit(emc, cores_for_chains = 3, cores_per_chain = 1,
           fileName = "data/rating-models.RData")
```

```{r, include = FALSE}
load("data/rating-models.RData")
```

## 5. Check the Fit

```{r}
summary(emc)
```

```{r, fig.alt = "Posterior parameter densities against true values for the MTLNR example", fig.height = 9}
plot_pars(emc, true_pars = p_vector, use_prior_lim = FALSE)
```

Posterior predictives can be summarized and plotted the same way as the data:

```{r, results = "hide"}
pp <- predict(emc)
```

```{r, fig.alt = "Posterior predictive response proportions and RT quantiles by confidence rating"}
plot_ratings(dat, pp, factors = "S")
```

## 6. The z-Transformed ROC

Recognition-memory confidence data are often summarized with a
z-transformed ROC: the cumulative "old" response probability (from the most
confident "old" response down to the most confident "new" response),
z-transformed, plotted for `old` (signal) trials against `new` (noise)
trials. `zroc()` computes it and `plot_zroc()` plots it, here with the
posterior predictive as a thin line per draw:

```{r, fig.alt = "z-transformed ROC for old versus new words, data and posterior predictive"}
plot_zroc(dat, "S", "new", post_predict = pp)
```

A straight zROC with unit slope (as above) is the classic signature of
equal-variance signal and noise distributions; here both accumulators share
the same `s`, so the model produces exactly that pattern. Letting `s` vary
with `lM` (an unequal-variance model) would tilt the slope away from 1.

## Next Steps

Other rating designs follow the same pattern: `n_ratings` in `MTLNR()` sets
the number of confidence levels, and the rating layer (`RR`, `rating_summary()`,
`plot_ratings()`, `zroc()`, `plot_zroc()`) is shared by the whole series of
rating models, not just the MTLNR.
