---
title: "Rating Models"
description: "Choice, response time and confidence ratings with the correlated multiple-threshold LNR"
author: "Andrew Heathcote"
output: rmarkdown::html_document
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{"Rating Models"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
rm(list = ls())
library(EMC2)
set.seed(30)
```

## Installation

This vignette uses the `dev-rating` branch:

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("ampl-psych/EMC2", ref = "dev-rating")
```

## Introduction

Rating tasks add a discrete confidence (or other ordered) rating to a choice
response time task: on every trial the data contain not just a choice `R` and
a response time `rt`, but also a rating `RR`, coded as an integer from 1
(the lowest rating) to `K` (the highest).

This vignette introduces the first rating model in *EMC2*, the correlated
multiple-threshold log-normal race (MTLNR). The MTLNR extends the log-normal
race (`?LNR`) to two *correlated* accumulators and reads the confidence
rating off the losing accumulator: at the moment the winner finishes, the
loser's state (as a proportion of its own threshold) is compared against
`K - 1` further thresholds to give one of `K` ratings. This is a natural
model for tasks such as old/new recognition memory with a confidence
judgment. For the full parameterization see `?MTLNR`.

> Reynolds, A., Kvam, P. D., Osth, A. F., & Heathcote, A. (2020). Correlated
> racing evidence accumulator models. *Journal of Mathematical Psychology*,
> *96*, Article 102331. https://doi.org/10.1016/j.jmp.2020.102331

The steps below are:

1. A minimal single-subject example with two ratings
2. A detailed single-subject example with three ratings: design, simulation,
   fitting, predictive checks and the z-transformed ROC
3. More than three ratings, and two current limitations
4. A hierarchical (multi-subject) example, extending the three-rating design

## 1. A Minimal Example: Two Ratings

The smallest rating design has `K = 2`: a single confidence criterion `c1`
per accumulator's losing evidence. We start with a single subject in a
generic two-choice task (`left`/`right`) with one shared criterion (`c1 ~ 1`,
not split by response), so the whole design has just six parameters.
`MTLNR()` is a factory: `MTLNR(n_ratings = K)` builds the model for `K`
ratings (`K = 3` is the default).

```{r, results = "hide"}
design_k2 <- design(
  factors = list(subjects = 1, S = c("left", "right")),
  Rlevels = c("left", "right"),
  matchfun = function(d) d$S == d$lR,
  formula = list(m ~ lM, s ~ 1, t0 ~ 1, rho ~ 1, c1 ~ 1),
  model = function() MTLNR(n_ratings = 2)
)
```

```{r}
sampled_pars(design_k2)
```

Simulating and fitting follow the same `design()` -> `make_data()` ->
`prior()` -> `make_emc()` -> `fit()` steps as any other *EMC2* model:

```{r, results = "hide"}
p_k2 <- sampled_pars(design_k2)
p_k2[] <- c(m = -.4, m_lMTRUE = -.7, s = log(.8), t0 = log(.3), rho = qnorm(.7), c1 = log(.5))
set.seed(20261001)
dat_k2 <- make_data(parameters = p_k2, design = design_k2, n_trials = 500)
```

```{r}
rating_summary(dat_k2)
```

```{r, results = "hide"}
prior_k2 <- prior(
  design = design_k2, type = "single",
  pmean = c(m = 0, m_lMTRUE = 0, s = log(.7), t0 = log(.25), rho = 0, c1 = log(.5)),
  psd = c(m = 1, m_lMTRUE = 1, s = .5, t0 = .4, rho = 1, c1 = .5)
)
emc_k2 <- make_emc(dat_k2, design_k2, prior_list = prior_k2, type = "single")
```

```{r, eval = FALSE}
emc_k2 <- fit(emc_k2, cores_for_chains = 3, cores_per_chain = 1,
              fileName = "data/rating-models-k2.RData")
```

```{r, include = FALSE}
# Loads every object cached for this vignette (k2, k3 and hierarchical); the
# later sections reload the same file after their own (also eval = FALSE)
# fit() calls, since their live design()/make_data() code above overwrites
# these names again for the narrative.
load("data/rating-models.RData")
```

```{r}
summary(emc_k2)
```

```{r, fig.alt = "Posterior parameter densities against true values for the two-rating example", fig.height = 7}
plot_pars(emc_k2, true_pars = p_k2, use_prior_lim = FALSE)
```

The rest of this vignette works through a fuller example (a response-specific
formula, a real-task narrative, predictive checks, and the zROC diagnostic);
the steps are the same ones used above.

## 2. A Detailed Example: Recognition Memory with Three Ratings

We simulate a two-choice recognition-memory task: participants see a word
that was either `new` or `old`, and respond `new` or `old` with a rating of
their confidence on a 3-point scale (`RR` = 1 low, 3 high).

### Specify the Design

As with the other race models (see `vignette("race-models")`), `matchfun`
flags whether the accumulator's response label matches the true word type,
giving the latent match factor `lM`; here it lets accuracy shift the mean log
finishing time. The two confidence criteria `c1` and `c2` are mapped `~ lR`
(one pair per response) since confidence thresholds belong to the *losing*
accumulator and so can differ between "new" and "old" responses, capturing a
response-specific bias in how confidence is reported.

```{r}
design_mtlnr <- design(
  factors = list(subjects = 1, S = c("new", "old")),
  Rlevels = c("new", "old"),
  matchfun = function(d) d$S == d$lR,
  formula = list(m ~ lM, s ~ 1, t0 ~ 1, rho ~ 1, c1 ~ lR, c2 ~ lR),
  model = MTLNR
)
sampled_pars(design_mtlnr)
```

### Rating Thresholds and Their Mapping

The criteria `c1`, `c2` are sampled on a log scale that guarantees the
underlying thresholds are ordered; `mapped_pars()` reports the natural-scale
thresholds `d1 < d2` (as proportions of the losing accumulator's own
threshold) alongside `rho`, mapped by response:

```{r}
p_vector <- sampled_pars(design_mtlnr)
p_vector[] <- c(m = -.5, m_lMTRUE = -.6, s = log(.8), t0 = log(.3), rho = qnorm(.75),
                c1 = log(.4), c1_lRold = .1, c2 = log(.6), c2_lRold = -.2)
mapped_pars(design_mtlnr, p_vector)
```

`rho` (here mapped through a scaled probit to lie in (-1, 1), `?MTLNR`) is
shared by both accumulators and is `r round(2 * pnorm(p_vector[["rho"]]) - 1, 2)`
for these parameter values: a strong positive trial-to-trial correlation
between the two finishing times.

### Simulate Data and Summarize by Choice and Rating

```{r, results = "hide"}
set.seed(20260930)
dat <- make_data(parameters = p_vector, design = design_mtlnr, n_trials = 1000)
head(dat)
```

`rating_summary()` gives, for each folded response (choice crossed with
rating), the response proportion and response-time quantiles:

```{r}
rating_summary(dat, factors = "S")
```

`plot_ratings()` plots the same information, response proportions on top and
RT quantiles below, split by word type:

```{r, fig.alt = "Response proportions and RT quantiles by confidence rating for new and old words"}
plot_ratings(dat, factors = "S")
```

### Set Priors and Fit

```{r, results = "hide"}
prior_mtlnr <- prior(
  design = design_mtlnr, type = "single",
  pmean = c(m = 0, m_lMTRUE = 0, s = log(.7), t0 = log(.25), rho = 0,
            c1 = log(.5), c1_lRold = 0, c2 = log(.5), c2_lRold = 0),
  psd = c(m = 1, m_lMTRUE = 1, s = .5, t0 = .4, rho = 1,
          c1 = .5, c1_lRold = .5, c2 = .5, c2_lRold = .5)
)
```

```{r, fig.alt = "Prior densities for the recognition-memory MTLNR example"}
plot(prior_mtlnr, N = 1e3)
```

```{r, results = "hide"}
emc <- make_emc(dat, design_mtlnr, prior_list = prior_mtlnr, type = "single")
```

```{r, eval = FALSE}
emc <- fit(emc, cores_for_chains = 3, cores_per_chain = 1,
           fileName = "data/rating-models.RData")
```

```{r, include = FALSE}
load("data/rating-models.RData")
```

### Check the Fit

```{r}
summary(emc)
```

```{r, fig.alt = "Posterior parameter densities against true values for the MTLNR example", fig.height = 9}
plot_pars(emc, true_pars = p_vector, use_prior_lim = FALSE)
```

Posterior predictives can be summarized and plotted the same way as the data:

```{r, results = "hide"}
pp <- predict(emc)
```

```{r, fig.alt = "Posterior predictive response proportions and RT quantiles by confidence rating"}
plot_ratings(dat, pp, factors = "S")
```

### The z-Transformed ROC

Recognition-memory confidence data are often summarized with a
z-transformed ROC: the cumulative "old" response probability (from the most
confident "old" response down to the most confident "new" response),
z-transformed, plotted for `old` (signal) trials against `new` (noise)
trials. `zroc()` computes it and `plot_zroc()` plots it, here with the
posterior predictive as a thin line per draw:

```{r, fig.alt = "z-transformed ROC for old versus new words, data and posterior predictive"}
plot_zroc(dat, "S", "new", post_predict = pp)
```

A straight zROC with unit slope (as above) is the classic signature of
equal-variance signal and noise distributions; here both accumulators share
the same `s`, so the model produces exactly that pattern. Letting `s` vary
with `lM` (an unequal-variance model) would tilt the slope away from 1.

## 3. More Than Three Ratings, and Two Current Limitations

**More ratings** are just a larger `n_ratings`: `MTLNR(n_ratings = K)` adds
one more criterion `c{K-1}` for every additional rating, each of which can be
mapped by any factor (here shared, `~ 1`, for brevity):

```{r}
design_k5 <- design(
  factors = list(subjects = 1, S = c("left", "right")),
  Rlevels = c("left", "right"),
  matchfun = function(d) d$S == d$lR,
  formula = list(m ~ lM, s ~ 1, t0 ~ 1, rho ~ 1, c1 ~ 1, c2 ~ 1, c3 ~ 1, c4 ~ 1),
  model = function() MTLNR(n_ratings = 5)
)
sampled_pars(design_k5)
```

**Limitation 1: the number of ratings must be the same for both responses.**
There is currently no way to give, say, 3 confidence levels to "new"
responses and 5 to "old" responses -- `n_ratings` must be a single integer,
shared by both choices:

```{r, error = TRUE}
MTLNR(n_ratings = c(3, 2))
```

**Limitation 2: MTLNR supports exactly two choices.** The confidence rating
is defined relative to the *losing* accumulator, and the correlation `rho` is
between exactly two log-normal finishing times, so `design()` refuses more
than two response levels:

```{r, error = TRUE}
design(factors = list(subjects = 1, S = c("a", "b", "c", "d")),
       Rlevels = c("a", "b", "c", "d"),
       matchfun = function(d) d$S == d$lR,
       formula = list(m ~ lM, s ~ 1, t0 ~ 1, rho ~ 1, c1 ~ 1),
       model = function() MTLNR(n_ratings = 2))
```

Both are scope limits of the current implementation, not the underlying
model; extending to unequal rating counts and to more than two choices (which
requires deciding which of the `> 1` losing accumulators' evidence defines
confidence) is planned future work.

## 4. A Hierarchical Example

The single-subject examples above use `type = "single"`. A hierarchical fit
estimates a group-level distribution over subjects along with each subject's
own parameters, using exactly the same design, `make_data()` and `fit()`
calls -- only the number of subjects, the prior, and `type` change. We extend
the three-rating recognition-memory design of Section 2 to 15 subjects, using
the same parameter values as the group means and adding modest between-subject
variability with `make_random_effects()`:

```{r, results = "hide"}
design_hier <- design(
  factors = list(subjects = 1:15, S = c("new", "old")),
  Rlevels = c("new", "old"),
  matchfun = function(d) d$S == d$lR,
  formula = list(m ~ lM, s ~ 1, t0 ~ 1, rho ~ 1, c1 ~ lR, c2 ~ lR),
  model = MTLNR
)

subj_sd <- c(m = .2, m_lMTRUE = .2, s = .15, t0 = .08, rho = .3,
             c1 = .2, c1_lRold = .2, c2 = .2, c2_lRold = .2)
set.seed(20261002)
subj_pars <- make_random_effects(design_hier, p_vector, n_subj = 15,
                                  covariances = diag(subj_sd[names(p_vector)]^2))
dat_hier <- make_data(subj_pars, design_hier, n_trials = 300)
```

```{r, results = "hide"}
prior_hier <- prior(
  design = design_hier, type = "standard", pmean = p_vector,
  psd = c(m = .3, m_lMTRUE = .3, s = .25, t0 = .15, rho = .5,
          c1 = .3, c1_lRold = .3, c2 = .3, c2_lRold = .3)
)
emc_hier <- make_emc(dat_hier, design_hier, prior_list = prior_hier, type = "standard")
```

`make_emc()` uses 3 chains by default; a hierarchical fit's per-iteration cost also scales with
the number of subjects, so it benefits from parallelizing *within* each chain across subjects too.
The call below assumes 9 cores are available (`cores_for_chains = 3` chains x `cores_per_chain = 3`
each); lower `cores_per_chain` (fitting will just take longer) if you have fewer.

```{r, eval = FALSE}
emc_hier <- fit(emc_hier, cores_for_chains = 3, cores_per_chain = 3,
                fileName = "data/rating-models-hier.RData")
```

```{r, include = FALSE}
load("data/rating-models.RData")
```

`summary()` on a hierarchical fit reports the group-level (`mu`, `Sigma`)
parameters by default:

```{r}
summary(emc_hier)
```

```{r, fig.alt = "Posterior group-level parameter densities against the group means for the hierarchical MTLNR example", fig.height = 9}
plot_pars(emc_hier, true_pars = p_vector, use_prior_lim = FALSE)
```

Everything from Section 2 (`rating_summary()`, `plot_ratings()`, `predict()`,
`zroc()`) works the same way on a hierarchical fit's data and predictives.

## Next Steps

Other rating designs follow the same pattern: `n_ratings` in `MTLNR()` sets
the number of confidence levels, and the rating layer (`RR`, `rating_summary()`,
`plot_ratings()`, `zroc()`, `plot_zroc()`) is shared by the whole series of
rating models, not just the MTLNR.
