---
title: "Response Models"
description: "Ordered and multinomial response models in EMC2"
author: "Niek Stevenson"
output: rmarkdown::html_document
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{"Response Models"}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo = FALSE}
rm(list = ls())
library(EMC2)
set.seed(21)
```

## Introduction

This vignette shows how to use the response-only models currently
implemented in *EMC2*. These models fall into two different response
geometries:

1.  Ordered response models: `ordered_probit()` and `ordered_logit()`
2.  Multinomial choice models: `multinomial_logit()` and
    `multinomial_probit()`

Ordered response models assume a single latent evidence variable and one
or more cutpoints. In the binary case this is the same general setup as
signal detection theory: a latent evidence value is compared against a
criterion. With more than two ordered responses, the single criterion
becomes a set of ordered criteria, as in rating or confidence versions
of Signal Detection Theory (SDT). `ordered_probit()` uses a Gaussian
latent noise distribution, which makes it the closest match to classical
Gaussian SDT. `ordered_logit()` uses the same threshold geometry but
with logistic noise.

Multinomial response models are different. They are not threshold models
on one evidence axis, but random-utility models over a set of
alternatives. `multinomial_logit()` is the standard multinomial-logit or
softmax model: each response has a latent utility and the response
probabilities are a softmax over those utilities. `multinomial_probit()`
uses the same utility geometry but with Gaussian noise instead of the
softmax/logit noise assumption.

The worked examples below use `ordered_probit()` and
`multinomial_logit()`. The other two models follow the same design
logic, but swap the latent noise distribution within the same geometry.

The workflow is the same as for the DDM and race models:

1.  Specify the design
2.  Inspect sampled parameters and their mapping to design cells
3.  Simulate data
4.  Define priors and create an `emc` object
5.  Fit the model
6.  Summarize posterior estimates and compare observed and posterior
    predictive choice proportions

```{r}
choice_summary <- function(data) {
  out <- as.data.frame(prop.table(table(data$S, data$R), 1))
  names(out) <- c("S", "R", "prob")
  out
}
```

## 1. Ordered Response Models

Ordered response models place responses on a single latent evidence axis
and use cutpoints to divide that axis into response categories. Here we
use a 3-category task with `low`, `mid`, and `high` stimuli and
responses. `location` depends on stimulus identity, `scale` is fixed to
1, and `cut ~ 1` estimates the standard `K - 1` thresholds.

Conceptually, this is the SDT-like response family in *EMC2*. In a
binary yes/no task there is one criterion. In a 3-category task there
are two ordered criteria. `location` shifts the latent evidence
distribution, and `cut` determines where the response boundaries lie.

```{r}
matchfun <- function(d) d$S == d$lR

design_ord <- design(
  Rlevels = c("low", "mid", "high"),
  factors = list(subjects = 1, S = c("low", "mid", "high")),
  formula = list(location ~ 0 + S, scale ~ 1, cut ~ 1),
  matchfun = matchfun,
  constants = c(scale = log(1)),
  model = ordered_probit
)
```

`sampled_pars()` shows the free parameters, and `mapped_pars()` shows
how they map to the latent design cells.

```{r}
sampled_pars(design_ord)
```

```{r}
p_vector_ord <- sampled_pars(design_ord)
p_vector_ord[] <- c(-1, 0, 1.2, -0.4, log(1.3))

mapped_pars(design_ord, p_vector_ord)
```

Now we simulate data from these parameters:

```{r, results = "hide"}
dat_ord <- make_data(parameters = p_vector_ord, design = design_ord, n_trials = 80)
```

For choice-only models it is often most useful to inspect response
proportions directly:

```{r}
choice_summary(dat_ord)
```

Next we define a prior and create the `emc` object.

```{r, results = "hide"}
prior_ord <- prior(
  design = design_ord,
  type = "single",
  pmean = c(
    location_Slow = -0.5,
    location_Smid = 0,
    location_Shigh = 0.5,
    cut_lRlow = -0.2,
    cut_lRmid = log(1.1)
  ),
  psd = c(
    location_Slow = 0.8,
    location_Smid = 0.8,
    location_Shigh = 0.8,
    cut_lRlow = 0.5,
    cut_lRmid = 0.3
  )
)

emc_ord <- make_emc(dat_ord, design_ord, prior_list = prior_ord, type = "single", n_chains = 2)
```

## 2. Fit the Ordered Model

The following call fits the model and saves the result:

```{r, eval = FALSE}
emc_ord <- fit(emc_ord, fileName = "data/response-models-ordered.RData")
```

```{r, include = FALSE}
load("data/response-models-ordered.RData")
emc_ord <- emc
rm(emc)
```

`summary()` reports posterior quantiles, `Rhat`, and ESS:

```{r}
summary(emc_ord)
```

`plot_pars()` compares the posterior to the generating values:

```{r, fig.alt = "Posterior parameter densities against true values for the ordered probit example"}
plot_pars(emc_ord, true_pars = p_vector_ord, use_prior_lim = FALSE)
```

Finally, `plot_fit_choice()` compares observed response probabilities to
posterior predictive intervals:

```{r}
plot_fit_choice(emc_ord, factors = "S", style = "prob", n_post = 20)
```

For ordered responses, cumulative fits can also be useful because they
reflect the threshold structure directly:

```{r}
plot_fit_choice(emc_ord, factors = "S", style = "cumulative", n_post = 20)
```

To switch from ordered probit to ordered logit, the design is identical
except for the model choice. That is, the latent threshold structure
stays the same, but the Gaussian link is replaced by a logistic one:

```{r, eval = FALSE}
design_ord_logit <- design(
  Rlevels = c("low", "mid", "high"),
  factors = list(subjects = 1, S = c("low", "mid", "high")),
  formula = list(location ~ 0 + S, scale ~ 1, cut ~ 1),
  matchfun = matchfun,
  constants = c(scale = log(1)),
  model = ordered_logit
)
```

## 3. Multinomial Response Models

Multinomial response models are for unordered choices among
alternatives. Here we use a three-choice task with stimuli `left`,
`right`, and `up`. The `matchfun` creates the latent match factor `lM`,
so the `utility` parameter can differ between the matching and
non-matching alternatives.

This is the softmax family in *EMC2*. `multinomial_logit()` maps the
latent utilities onto response probabilities through the
multinomial-logit link, i.e. through a softmax over alternatives. Unlike
the ordered models, there is no single latent evidence axis and no
ordered criteria.

Only utility differences are identified in the multinomial models. A
common utility shift across all alternatives leaves the softmax
probabilities unchanged, so the utility intercept must be fixed rather
than estimated. Here we anchor it at `0` and estimate only the
matching-versus-non-matching contrast.

```{r}
ADmat <- matrix(c(-1/2, 1/2), ncol = 1, dimnames = list(NULL, "d"))

design_mnl <- design(
  Rlevels = c("left", "right", "up"),
  factors = list(subjects = 1, S = c("left", "right", "up")),
  formula = list(utility ~ lM),
  contrasts = list(utility = list(lM = ADmat)),
  matchfun = matchfun,
  constants = c(utility = 0),
  model = multinomial_logit
)
```

Again we inspect the sampled parameters and their mapping:

```{r}
sampled_pars(design_mnl)
```

```{r}
p_vector_mnl <- sampled_pars(design_mnl)
p_vector_mnl[] <- 2

mapped_pars(design_mnl, p_vector_mnl)
```

We simulate a dataset from this multinomial-logit model:

```{r, results = "hide"}
dat_mnl <- make_data(parameters = p_vector_mnl, design = design_mnl, n_trials = 80)
```

And inspect the observed choice proportions:

```{r}
choice_summary(dat_mnl)
```

Now we specify priors and create the `emc` object.

```{r, results = "hide"}
prior_mnl <- prior(
  design = design_mnl,
  type = "single",
  pmean = c(utility_lMd = 1.5),
  psd = c(utility_lMd = 0.7)
)

emc_mnl <- make_emc(dat_mnl, design_mnl, prior_list = prior_mnl, type = "single", n_chains = 2)
```

## 4. Fit the Multinomial Model

```{r, eval = FALSE}
emc_mnl <- fit(emc_mnl, fileName = "data/response-models-multinomial.RData")
```

```{r, include = FALSE}
load("data/response-models-multinomial.RData")
emc_mnl <- emc
rm(emc)
```

```{r}
summary(emc_mnl)
```

```{r, fig.alt = "Posterior parameter densities against true values for the multinomial logit example"}
plot_pars(emc_mnl, true_pars = p_vector_mnl, use_prior_lim = FALSE)
```

```{r}
plot_fit_choice(emc_mnl, factors = "S", style = "prob", n_post = 20)
```

To use Gaussian utility noise instead, the design is the same and only
the model changes:

```{r, eval = FALSE}
design_mnp <- design(
  Rlevels = c("left", "right", "up"),
  factors = list(subjects = 1, S = c("left", "right", "up")),
  formula = list(utility ~ lM),
  contrasts = list(utility = list(lM = ADmat)),
  matchfun = matchfun,
  constants = c(utility = 0),
  model = multinomial_probit
)
```

The core workflow is therefore the same for all four response models:
`design()` -\> `prior()` -\> `make_emc()` -\> `fit()` -\> posterior
predictive comparison of choice proportions.

The main conceptual map is:

-   `ordered_probit()`: SDT-like ordered responses with Gaussian latent
    noise
-   `ordered_logit()`: the same threshold model with logistic latent
    noise
-   `multinomial_logit()`: unordered choice with a softmax /
    multinomial-logit link
-   `multinomial_probit()`: unordered choice with Gaussian utility noise
