---
title: "Sampler Validity"
description: "Why EMC2's particle sampler was replaced, what the current sampler does, and the evidence that it samples the right posterior, including against Stan"
author: "Andrew Heathcote"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{"Sampler Validity"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
rm(list = ls())
library(EMC2)
set.seed(1)
```

This vignette is for users who want to know what EMC2's sampler does and how far to trust it, and for whoever
maintains it. It explains why the previous particle sampler was not a valid MCMC sampler, describes the current
one, and gives the evidence that the current one samples the right posterior: exact references, Stan's NUTS on
the same posteriors, and ordinary fits. The previous sampler is still in the package, so every comparison can be
rerun with one switch:

```{r, eval = FALSE}
options(emc.sampler = "legacy")   # the previous particle step, tuning and sample stage
options(emc.sampler = NULL)       # the current sampler
```

Throughout, **excess/k** = (mean sample-stage deviance - deviance at the posterior mode) / number of parameters,
which is 1 for a normal posterior; **SD ratio** = posterior SD / Laplace SD (from the numerical Hessian at the
mode); **move rate** = the fraction of iterations whose value changed.

## 1. Why the previous sampler was replaced

**The particle step.** Each iteration, for each subject, `new_particle()` draws a batch of particles from a mixture
of multivariate normals, keeps the chain's current value as particle 1, gives every particle the weight

    w(z) = exp( loglik(z) + logprior(z) - log q(z) )

with `q` the mixture density, and selects one particle in proportion to `w`. With `q` fixed this is iterated
sampling-importance-resampling (i-SIR; Andrieu, Lee & Vihola, 2018), and it leaves the posterior invariant. In the
previous sampler `q` is not fixed: in every stage some components are centred on the current value (the "random
walk" `N(x, epsilon^2 chains_var)`), and `q` is evaluated at the current value itself, at the *peak* of those
components. The current value's weight is deflated by a factor that depends on where the chain is, and the kernel
is no longer posterior-invariant.

With one candidate `z ~ N(x, S)` the move probability is `pi(z) N(x|x,S) / (pi(z) N(x|x,S) + pi(x) N(z|x,S))`. Since
`N(x|x,S) >= N(z|x,S)`, this exceeds the Barker probability `pi(z)/(pi(z) + pi(x))`, by more the further `z` is
from `x`. Detailed balance fails, and more particles do not help when the importance weights are degenerate, which
is the high-dimensional, concentrated-posterior case.

**The sample stage.** The previous sample stage re-estimates its proposals (`chains_var`, `chains_mu`, `eff_mu`,
`eff_var`) from the chain's last 250 draws at every step, and keeps adapting its step size and mixing weights. A
chain whose kernel depends on its recent past is not a Markov chain with the posterior as its stationary
distribution. Where the posterior's scale does not move this is negligible. In a hierarchical funnel it is not: on
a conjugate model with a known posterior (25 parameters x 20 subjects), chains started at the truth sink into the
neck. For the weakly identified parameters the group SD has a median of 0.004, against 0.16 exact, with 94% of the
draws in the neck against 26%.

**Consequences.**

- With 9 parameters the bias is a few percent, which is why simulation-based calibration on small models does not
  show it.
- With 25-35 parameters, converged single-subject fits have excess/k of 1.5-1.9 and SD ratios of 1.25-1.4 (up to
  1.67).
- With 29-35 parameters and at least 2000 trials, chains stick: move rate .01, Rhat 1.2-2, posterior means up to
  18 posterior SDs from the truth.
- The previous tuning made sticking worse:
  - a `diag(.5)` fallback for an unusable chain covariance, 35 times too wide for a posterior SD of .02;
  - step-size floors of .1, .4 and .6;
  - shrinking the independence components on low acceptance, although for them low acceptance means too
    *narrow*.
- On the hard cases of section 5, the previous sampler reports Rhat 1.01-1.09 while being off by up to 46
  standard errors from Stan's NUTS on the same posterior.

## 2. The bias on a Gaussian toy

The function below is the previous sample-stage step: the same particle numbers (`50 sqrt(d)`), the same mixture
weights (.05 prior, .3 random walk at the current value, .3 at a fixed mean, .35 "efficient" proposal), and a target
that is the likelihood `N(0, I_d)` times a `N(0, 100 I_d)` prior. `fix = TRUE` is the ensemble move of section 3.
Ten seconds per row.

```{r}
library(mvtnorm)
step <- function(x, d, eps, n, fix = FALSE) {
  z0 <- rep(0, d); I <- diag(d); P <- diag(100, d)
  # component 2 is the random walk: centred on x (previous step) or on an auxiliary centre c ~ N(x, S)
  S2 <- eps^2 * I
  c2 <- if (fix) drop(rmvnorm(1, x, S2)) else x
  Mus <- list(z0, c2, z0, z0); Ss <- list(P, S2, eps^2 * I, eps^2 * I); mix <- c(.05, .3, .3, .35)
  pn <- pmax(1, rmultinom(1, n, mix))
  props <- rbind(x, do.call(rbind, lapply(1:4, function(j) rmvnorm(pn[j], Mus[[j]], Ss[[j]]))))
  lpost <- dmvnorm(props, z0, I, log = TRUE) + dmvnorm(props, z0, P, log = TRUE)
  lq <- log(Reduce(`+`, lapply(1:4, function(j) mix[j] * dmvnorm(props, Mus[[j]], Ss[[j]]))))
  extra <- if (fix) dmvnorm(props, c2, S2, log = TRUE) else 0   # target factor N(z | c, S) of the ensemble move
  w <- exp(lpost + extra - lq - max(lpost + extra - lq))
  props[sample(nrow(props), 1, prob = w), ]
}
toy <- function(d, eps, fix = FALSE, iters = 1500, n = round(50 * sqrt(d))) {
  x <- rnorm(d); X <- matrix(NA, iters, d)
  for (i in 1:(300 + iters)) { x <- step(x, d, eps, n, fix); if (i > 300) X[i - 300, ] <- x }
  c(d = d, eps = eps, fix = fix, var_over_true = round(mean(apply(X, 2, var)) / (1 / (1 + 1/100)), 2))
}
rbind(toy(25, .6), toy(25, .3), toy(25, 2), toy(9, .6), toy(35, .6), toy(25, .6, fix = TRUE))
```

The chain "moves" every iteration, but its variance follows `eps`, not the posterior. Narrower components pull it
down and wider ones push it up, more so as `d` grows. In longer runs the previous step gives 0.65 times the true
variance at `eps = .6`, 0.20 at .3 and 1.47 at 2; the ensemble move gives 0.98-1.02 everywhere.

## 3. The sampler

### The particle step

`preburn` and `burn`, whose draws are discarded, use the previous step as a search: it is a fast stochastic
hill-climb from prior-scale starting values, and a valid random-walk kernel is far slower at that. `adapt` and
`sample`, whose draws build the proposals and the posterior, use an exact kernel. Each iteration runs ONE of two
kernels, chosen with the mixture weights:

- *Local kernel* (the components centred on the current value): an ensemble move (Tjelmeland, 2004; Neal, 2011).
  - Draw an auxiliary centre `c ~ N(x, S)`, and draw the particles from `N(c, S)`.
  - Weight every particle, the current value included, by `pi(z) N(z | c, S) / q(z)`, where `q` is the proposal
    density, which given `c` does not depend on the state.
  - The step targets `pi(x) N(c | x, S)`, whose `x`-marginal is `pi`, so it is exact.
- *Global kernel* (the prior, `N(chains_mu, chains_var)` or its conditional replacement below, `N(eff_mu,
  eff_var)`): i-SIR, which is exact because none of these depends on the current value.

Keeping the two kernels separate, rather than one batch with the ensemble correction, gives 4-6 times the
effective sample size per iteration on the toy.

### One fixed kernel in the sample stage

The sample stage runs one kernel, which is never changed once draws are kept:

- `adapt` runs until the draws the kernel will be built from have converged. For models with a group level it
  waits until, over its last 250 iterations, the largest Rhat of any subject's parameter is below 1.2. That is
  checked from 250 iterations on, and adapt gives up after three checks without improvement or at 1000 iterations.
  `options(emc.adapt_converge = FALSE)` switches this off and keeps the older `min_unique` rule.
- The sample-stage proposals are then built from those draws and stored in `emc[[1]]$sample_kernel`. Every step of
  the sample stage, and every later call that adds samples, reuses them.
- A final 100-iteration tail of `adapt` runs that kernel and tunes it (`tune_sample_kernel()`):
  - the local step size is set to its average over the tail;
  - the number of particles is set where the average effective sample size meets its target
    (`2.5 sqrt(p)`, at least 25 particles).
- In `sample` nothing adapts.

### Proposals that follow the group level

The subject step conditions on the group level `(mu, Sigma)`, so its proposals may depend on it. When a group
variance collapses, the subject's conditional posterior becomes much narrower in those coordinates only, and no
single step size fits. For every type with a group level:

- `create_lik_prec()` estimates each subject's likelihood precision `L` (minus the Hessian of the log-likelihood),
  and the linear term `l` of the same quadratic approximation, at the mean of the subject's recent draws:
  - *by central differences* (`lik_precision()`), with each parameter's step searched for so that the
    log-likelihood drops by about 1. The search brackets the step and bisects in log scale. A parameter whose step
    never settles, as at a likelihood floored at a model bound such as the DDM's `sv` and `SZ`, gets no precision
    and no gradient rather than a cliff read as a huge curvature;
  - *from the draws* (`type = "standard"`): the draws' inverse covariance minus the window's mean group precision,
    in the directions where that is clearly positive, with the finite differences (capped at what the draws allow)
    elsewhere. One point's curvature is too narrow for a skewed posterior. The subtraction is right only if the
    group precision is about constant over the window, so this version is used only where the largest coefficient
    of variation of the group precision is below .5. In a funnel it is not, and the finite differences are kept.
- The local kernel's covariance is `epsilon^2 (L + Sigma^-1)^-1` at the current `Sigma`, so the step size is a
  relative scale that no longer has to track the group level.
- The global kernel's `N(chains_mu, chains_var)` component is the Gaussian approximation of the subject's
  conditional posterior at the current group level, `N((L + Sigma^-1)^-1 (l + Sigma^-1 mu), (L + Sigma^-1)^-1)`.
- The local kernel keeps at least a quarter of the sample-stage iterations.

`type = "single"` has no group level and uses the chain covariance.

**Tuning.**

- The local step size targets 3% of particles beating the current value; on the toy that is where effective
  sample size per iteration peaks in 9, 25 and 35 dimensions.
- It is clamped to [.01, 20] and updated only in iterations that used the local kernel, from that iteration's own
  rate.
- Independence components are used at the scale of their estimated covariance (epsilon 1).
- An unusable chain covariance is replaced by the mean of the other subjects' covariances, else the chain's
  previous one, else 0.1 times the prior variance.

### The interweaving sweep (`type = "standard"`)

With few trials per weakly identified parameter, the alternation "subjects given the group level, group level given
the subjects" mixes very slowly in the funnel, even with an exact subject step. On the conjugate model, one
effective draw of a weak parameter's log group SD takes about 500 iterations. The remedy is interweaving, or
non-centred reparameterisation (Yu & Meng, 2011; Papaspiliopoulos, Roberts & Sköld, 2007): after the group step,
`scale_move_standard()` moves in a parameterisation in which the group scale and the subjects' deviations are not
conditionally dependent. For each parameter `j`:

- a **scale move** proposes a log-scale change `delta` and sets `alpha'_sj = mu_j + f (alpha_sj - mu_j)` for every
  subject (`f = exp(delta)`), `Sigma' = D Sigma D` with `D = diag(1, .., f, .., 1)`, and the Huang-Wand auxiliary
  `a'_j = a_j / f^2`, so that the prior's own hierarchy moves along (Huang & Wand, 2013). Its prior, group-density
  and Jacobian part of the log acceptance ratio is `delta - (exp(2 delta) - 1) / (A_j^2 a_j)`;
- a **location move** shifts `mu_j` and every subject's `alpha_sj` together by `e`. It runs only without a
  group-level design.

The moves are proposed on each subject's quadratic likelihood surrogate (`L`, `l` above), in random blocks of
parameters. Each block's combined proposal is then accepted or rejected against the subjects' exact likelihoods,
in one more likelihood evaluation per subject (delayed acceptance; Christen & Fox, 2005). The chain therefore
targets the exact posterior whatever the surrogate's quality.

Tuning, in `adapt` only; everything is frozen for `sample`:

- *Step sizes* adapt by Robbins-Monro on the *realised* acceptance: the surrogate accepted and the block passed the
  exact check. The target is `.3 x max(r, .35)`, where `r` is the running block acceptance (a move shares its
  block's fate).
- *Starting steps* are each move's conditional scale on the surrogate, where that is below the default.
- *The number of blocks* adapts to keep the block acceptance in [.35, .75].
- *A gate* drops, for the rest of the fit, any move whose expected squared jump per sweep is under a tenth of the
  group-level Gibbs step's own, once the sweep is split into one block per parameter. It is decided from 300
  sweeps on. With hundreds of subjects the surrogate's error adds up, so moves of well-identified parameters pass
  only at tiny steps while each block costs one likelihood per subject.
- *The exact check* is timed in-process and forked on its first calls, and the faster way is kept.

The state is in `attr(samples, "scale_move")`. `options(emc.scale_move = FALSE)` turns the sweep off; the legacy
sampler never runs it.

### Rhat and the stop rules

`gd_summary()`, `summary()`, the Rhat on plots and `fit()`'s stop rules all use one statistic, `split_rhat()`:
each chain is halved, and Gelman & Rubin's (1992) potential scale reduction factor is taken over the half-chains
(Gelman et al., 2013), with no degrees-of-freedom correction and no transform. coda's `gelman.diag()` point
estimate multiplies by a correction that grows when the chains' variances differ, which is what one chain
visiting a skewed posterior's shoulder does. Such a subject reads 1.05 on the split-Rhat and 1.15-1.2 on coda's
estimate, and with thousands of subject-level entries the largest of them decides a fit.

- `burn` stops when the *mean* Rhat of the selection is below 1.1; `sample` stops when the *largest* is, with at
  least 1000 draws kept, checked every `step_size` draws for at most `max_tries` checks.
- On a failed check, the first third of the stage's draws is dropped if the largest Rhat is lower without it.
  This removes chains still arriving at the posterior. Chains that drift together have a low Rhat, so the
  criterion alone would keep such a transient (section 5).
- A check is one pass over the stored draws: half a second with 518 subjects.
- `stop_criteria$gd_quantile` (e.g. `.99`, not the default) judges the subject level by that quantile of its
  Rhats instead of the largest. Every other selected parameter, and the discard, still use the largest. With it,
  the criterion does not tighten as subjects are added.

### Other behaviour

- If a draw of the group covariance is numerically singular, the group step is re-drawn up to 3 times before the
  chain errors (`on_singular`; `list(max_retries = 0)` errors at once, and `on_exhausted = "carry_forward"` keeps
  the previous group parameters).
- On Windows the chains run in separate R processes; the `emc.*` options are copied to them.

## 4. The bias in an EMC2 fit

`vignettes/data/build_sampler_validity.R` simulates one subject (about 500 trials) from a 25-parameter LNR
(`m, s ~ 0 + NWS:WF:lM, t0 ~ 1`, with parameter values from a real participant). It fits the data with both
samplers using `fit()` defaults, and computes the Laplace SDs. It takes two minutes on three cores; the results
are cached.

```{r}
load("data/sampler-validity.RData")
tab <- t(sapply(summaries, function(s) c(
  `excess/k` = round(s$excess / s$k, 2),
  `SD/Laplace (median)` = round(median(s$pars$post_sd / s$pars$laplace_sd), 2),
  `SD/Laplace (min)` = round(min(s$pars$post_sd / s$pars$laplace_sd), 2),
  `SD/Laplace (max)` = round(max(s$pars$post_sd / s$pars$laplace_sd), 2),
  `max |z|` = round(max(abs(s$pars$z)), 1), `move rate` = round(s$move_rate, 2),
  `max Rhat` = round(s$max_rhat, 3), iterations = sum(s$iters))))
tab
```

```{r, fig.width = 7, fig.height = 4}
r <- sapply(summaries, function(s) s$pars$post_sd / s$pars$laplace_sd)
plot(r[, "legacy"], type = "h", lwd = 6, col = "grey60", ylim = c(0, max(r) * 1.05), xaxt = "n",
     xlab = "parameter", ylab = "posterior SD / Laplace SD", main = "25-parameter LNR, ~500 trials")
points(seq_len(nrow(r)) + .25, r[, "current"], type = "h", lwd = 6, col = "#0072B2")
abline(h = 1, lty = 2); axis(1, at = seq_len(nrow(r)), labels = FALSE)
legend("topleft", c("legacy sampler", "current sampler"), fill = c("grey60", "#0072B2"), bty = "n")
```

The legacy sampler reports Rhat of about 1 and a move rate of 1.00, while every posterior SD is 1.2-1.6 times too
wide. The current sampler's SDs match the Laplace SDs, with an honest move rate of about .8. The tuning state is in
`attr(emc[[1]]$samples, "pm_settings")`:

```{r}
sapply(summaries, function(s) round(s$epsilon, 2))
```

Under the legacy sampler the independence components end at a scale above 1, wider than the posterior, which is
where the over-dispersion comes from: the chain follows them. Under the current sampler they are fixed at 1, and
only the local step is scaled.

## 5. The evidence

The scripts behind this section are available from the package maintainers.

### Exact references

- *Single subjects:* a grid of 9, 25 and 35 parameters x 500, 2000 and 8000 trials. With the current sampler,
  excess/k is 0.90-1.00, the median SD ratio is 0.98-1.03 and Rhat is at most 1.09 in every cell. With the legacy
  sampler excess/k is 1.5-1.9 and the SD ratios are 1.25-1.4 from 25 parameters up, and fits with at least 2000
  trials and 29 or more parameters stick.
- *A conjugate hierarchical model with a known posterior:* 25 parameters x 20 subjects, 12 of them weakly
  identified, with the exact posterior from a Gibbs sampler of the full conditionals. Starting both from the truth
  and from a collapsed state:
  - the current sampler reproduces the group-SD quantiles (1-95%) to three decimals in every identifiability
    class, and the share of draws in the funnel's neck (.25 against .24-.26);
  - subject-level SD ratios are 0.997-1.003;
  - the autocorrelation time of a weak parameter's log group SD is 10-16 iterations, against about 500 without the
    sweep;
  - the median first escape from a collapsed start is 28 iterations.

### Against Stan's NUTS

Nine hard cases were run in Stan (Carpenter et al., 2017; Hoffman & Gelman, 2014) on exactly EMC2's posterior: the
same data, priors (with the Huang-Wand auxiliaries integrated out in closed form), bounds and per-trial likelihood
floor. The Stan group level was checked against the exact reference above, and each Stan likelihood against
`calc_ll_manager()`. EMC2 was run on the same `emc` objects, with five `fit()`s with defaults and 3-15 long runs
(5000-10000 sample draws, no stop rule) per case, 3 chains each.

"Agrees" means every group mean, log group SD, group correlation and the parameters of three subjects are within
Monte-Carlo error of Stan. ESS ratio is EMC2's effective draws per core-hour over Stan's, for the median
group-level quantity and the worst one.

| case | Stan's verdict | EMC2 long runs against Stan | ESS per core-hour, EMC2 / Stan: median; worst | EMC2's verdict |
|---|---|---|---|---|
| LNR funnel, 20 subjects x 200 trials, 25 parameters | stiff but sampled (~1000 leapfrogs per draw) | agrees (6 x 10000) | 2.6; 0.6 | right but slow; partly the posterior |
| LNR funnel, 8 subjects x 500 trials | no problem | agrees (5 x 10000) | 0.52; 0.012 | right but slow |
| simulated DDM, `sv` / `SZ` near the floor | no problem | agrees (5 x 10000) | 7.6; 0.15 | right but slow |
| simulated DDM, `sv` / `SZ` inside | no problem, costly per gradient | agrees (5 x 5000) | 44; 4.1 | right but slow; cheaper than NUTS |
| forstmann, `sv` | no problem; the `sv` plateau is real and crossed | agrees (5 x 10000) | 15; 0.020 | right but slow |
| forstmann, `sv` + `SZ` | no problem; same plateau | agrees within a wide error (15 x 10000) | 97; 0.54 | right but very slow |
| Eisenberg Simon (RDM, 518 subjects) | no problem | agrees (3 x 5000) | 0.51; 0.25 | right but slow |
| simulated RDM, 300 subjects | at its tree-depth limit | agrees (3 x 5000) | 0.97; 0.33 | right but slow; NUTS works hard too |
| simulated RDM, 300 subjects, correlated funnel | 32-50 effective draws of the funnel SDs | agrees within a very wide error (3 x 5000) | 0.54; 0.011 | hard for both |

**Monte-Carlo error.** EMC2's chains carry offsets that last longer than a run, which the within-chain ESS cannot
see, so a single run can look 5-8 ESS-based standard errors from Stan. The comparison therefore uses:

- the larger of the ESS-based and a chain-batch standard error;
- the share of EMC2's draws beyond Stan's 5% and 95% points;
- several independent long runs per case.

What settles whether a difference is EMC2's error or a different posterior is EMC2 against itself: z-scores of the
`fit()` runs against the long runs, built the same way. SD of the z-scores over every subject parameter:

| case (subject parameters) | `fit()` vs Stan | long vs Stan | `fit()` vs long |
|---|---|---|---|
| LNR funnel, 20 subjects (500) | 1.21 | 1.48 | 1.36 |
| LNR funnel, 8 subjects (200) | 1.85 | 1.31 | 1.57 |
| DDM near the floor (90) | 0.91 | 0.84 | 0.96 |
| DDM inside (90) | 1.13 | 1.18 | 1.07 |
| forstmann `sv` (152) | 1.41 | 1.29 | 1.76 |
| forstmann `sv` + `SZ` (171) | 0.93 | 1.48 | 1.30 |
| Eisenberg (4144) | 1.19 | 1.45 | 1.40 |
| RDM, 300 subjects (1800) | 0.99 | 1.46 | 1.32 |
| RDM, correlated funnel (2400) | 1.46 | 1.38 | 1.22 |

The z-scores are 1.2-1.5 times too wide, and just as wide between EMC2's own runs: EMC2's ESS understates its
Monte-Carlo error by about 1.5-2 in variance, and the posterior is the same. With chain-batch errors, the largest
group-level |z| per case is 2.2-3.3 over 49-425 quantities.

**Where EMC2 is slow.** On six of the nine cases NUTS has no difficulty and EMC2 needs long runs. Per core-hour EMC2
is 8-97 times cheaper than NUTS on the typical quantity of the DDMs, whose gradients are expensive. On the slowest
quantity of eight of the nine cases it is 1.7-90 times slower: funnel necks, plateaus at a parameter's bound,
ridges along weakly identified combinations. In long runs a chain enters a neck or a plateau and stays 50-5000
iterations; NUTS's longest stay beyond its own 5% point is 2-8 draws. The clearest case is forstmann with `sv` +
`SZ`:

- a chain occasionally moves onto a region where `SZ` sits at its lower bound for every subject, and stays there
  for thousands of iterations;
- the region is real: the likelihood there is as high as at Stan's mode, and Stan visits it too (0.8% of its
  draws);
- EMC2's share is 4.6%, with a run-level SE of 2.9%. That is within Monte-Carlo error, but 45 chains x 10000 draws
  don't pin that group level, where NUTS needs four chains of 1000.

**The legacy sampler is wrong, with good Rhat.** `options(emc.sampler = "legacy")`, three `fit()` runs each:

| | alpha Rhat at stop | group means / log SDs / correlations: largest \|z\| | subject parameters beyond \|z\| = 4 | SD ratios |
|---|---|---|---|---|
| Eisenberg | 1.01-1.02 | 26 / 46 / 26 | 2966 of 4144 | 0.53-2.46 |
| forstmann, `sv` + `SZ` | 1.04-1.09 | 9.2 / 25 / 16 | 50 of 171 | 0.33-1.08 |

### Ordinary fits: forstmann with five standard models

Hierarchical fits of `forstmann` (19 subjects) with LNR, LBA, RDM, the Wiener DDM (`sv`, `SZ`, `st0` at 0) and the
full DDM (all three estimated):

- speed emphasis affects the threshold only (for LNR, the log mean);
- there is no rate bias, and response bias enters through `lR` (`Z` for the DDMs);
- the LBA's `sv` and the RDM's `s` vary with `lM`, with the mismatch level fixed at 1, and the RDM has no
  start-point variability.

Every fit is `fit()` with all defaults, five runs per model and sampler, on one 48-core machine. Times are medians.

| model | converged by the default rule, current / legacy | minutes to stop, current / legacy | s per 100 sample iterations, current / legacy | posteriors |
|---|---|---|---|---|
| LNR | 5/5 / 5/5 | 1.9 / 2.1 | 5.4 / 8.2 | the same |
| RDM | 5/5 / 5/5 | 2.7 / 3.6 | 6.7 / 15.8 | the same |
| Wiener DDM | 5/5 / 5/5 | 2.3 / 2.3 | 6.0 / 9.7 | the same |
| LBA | 2/5 / 5/5 | 7.4 / 5.7 | 15.8 / 23.9 | legacy too narrow |
| full DDM | 4/5 / 5/5 | 55.5 / 58.6 | 96 / 201 | legacy differs in `SZ` |

- *Time to convergence is about the same.* Per sample iteration the current sampler is 1.5-2.4 times faster, because
  one fixed kernel needs fewer particles than continual re-tuning. Its converged adapt stage takes most of that
  back.
- *LNR, RDM and the Wiener DDM give the same posteriors:* the largest group-level shift is .03 posterior SDs.
- *LBA.* Against four long current-sampler runs (5000 draws, 7 minutes each, which agree with each other):
  - the current sampler's default fits agree, including those that do not meet the stop rule;
  - the legacy sampler's posterior SDs are 20-33% too small for 18 of 171 subject parameters (`A`, `t0` and `B`
    of five subjects);
  - the current sampler misses the rule within `fit()`'s limit in 3 of 5 runs. Each time it is one subject whose
    parameters sit on strong posterior trade-offs (`v` with `v_lMd` -.97, with `sv` .95; `B` with `t0` -.79) and
    drift along that ridge for a few hundred iterations. More draws are the remedy.
- *Full DDM.* The two samplers differ in `SZ`: the group means are .46 posterior SD apart, and the legacy sampler's
  spread of the group SD of `SZ` is 2.5 times narrower. That is the signature of the Stan comparison above, where
  the legacy sampler is the one that is wrong.

### The stop rule

With the sample-stage kernel fixed, a stop rule only decides where to stop and which draws to keep, so rules can be
compared exactly on runs that kept every draw. 41 candidate rules were compared on 40 runs of six data sets
(Eisenberg, two LNR funnel models, forstmann's full DDM, two simulated 300-subject RDMs). Each run was also
replayed with its own burn and adapt draws put in front as a real transient.

- *The statistic.* On the split-Rhat, the rule stops on its criterion in 18 of 20 Eisenberg runs, after a median
  1380 draws. On coda's estimate it stops in 3 of 20, at the 2000-draw limit. The retained draws are as accurate as
  a window of the same length from the end of the run.
- *The discard is needed.* Without it a transient stays in, and the fit still passes: chains that drift together
  have a split-Rhat below 1.1. On Eisenberg the error of the retained draws is 1.6 (subject level) to 2.5 (group
  level) times a stationary window's, and 3-9 times on the small funnel models. With the discard, every case is back
  to a stationary window.
- *Nothing cheaper is as safe.* A margin, a cap at half the draws and a fixed warm-up fraction each remove a long
  transient less well. A test for drift (batch means, or Geweke's diagnostic pooled over chains) cannot tell a
  transient from slow mixing, and fires at 24-81% of the checks of slowly mixing fits.
- *In `fit()`*, 3 of 5 Eisenberg runs stop on the criterion, in 37 minutes, with every group-level quantity within
  2.3 Monte-Carlo SEs of Stan.
- *`gd_quantile = .99`* loses nothing on the transient test, and stops the small funnel models at 1200-1400 draws
  (4-5 of 5 runs) instead of at the limit. It lets 1-5 subject-level entries stand at 1.15-1.35 at the stop.

## 6. What is still open

- **Slow directions.** The current sampler is right but slow along group-level ridges, in funnel necks, on
  plateaus at a bound (forstmann's `SZ`), and along strong within-subject trade-offs (LBA); more draws are the
  remedy. Moves along these directions would be the sampler's remedy. Joint group-level moves along the slowest
  directions (shift, rescale, shear) are exact, and halve a simulated ridge's autocorrelation time, but they cost
  as much time as they save on real data, so they are not in the package.
- **Optimistic ESS.** EMC2's ESS understates Monte-Carlo error by about 1.5-2 in variance, and
  `stop_criteria$min_es` inherits that.
- **Stop-rule details.** `max_tries` does not bind until 1000 draws are kept. Whether `gd_quantile = .99` should be
  the default needs more designs.
- **The sample kernel's step size on small models.** The local step size is averaged over the 100-iteration tail
  only after 20 warm-up uses, and needs 10 more. On a 4-subject model many subject-chains keep their last value
  instead; on forstmann's LBA 96-100% are averaged. This costs efficiency, not correctness.
- **Singular group covariances.** Whether 3 re-draws are always enough is not known.
- **Scope of the sweep.** It runs for `type = "standard"` only, and its location move only without a group-level
  design. There is no sweep in `burn`, where `init()` starts every chain near the funnel's neck. Dispersed
  starting values have not been tried.
- **Found while matching Stan's target, outside the sampler:**
  - the RDM's Wald survivor `pigt0()` drops the loser's term when `2 B v / s^2 > 709.8`, a region the sampler can
    reach;
  - the DDM's precision constant `DDM_EPS = 5e-3` changes no posterior here, and `1e-6` costs nothing measurable at
    the level of a `fit()`;
  - models with nuisance parameters cannot currently be fitted end to end.

## References

Andrieu, C., Lee, A., & Vihola, M. (2018). Uniform ergodicity of the iterated conditional SMC and geometric
ergodicity of particle Gibbs samplers. Bernoulli, 24(2), 842-872.

Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M., Guo, J., Li, P., &
Riddell, A. (2017). Stan: A probabilistic programming language. Journal of Statistical Software, 76(1), 1-32.

Christen, J. A., & Fox, C. (2005). Markov chain Monte Carlo using an approximation. Journal of Computational and
Graphical Statistics, 14(4), 795-810.

Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., & Rubin, D. B. (2013). Bayesian data analysis
(3rd ed.). CRC Press.

Gelman, A., & Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. Statistical
Science, 7(4), 457-472.

Hoffman, M. D., & Gelman, A. (2014). The No-U-Turn sampler: Adaptively setting path lengths in Hamiltonian Monte
Carlo. Journal of Machine Learning Research, 15, 1593-1623.

Huang, A., & Wand, M. P. (2013). Simple marginally noninformative prior distributions for covariance matrices.
Bayesian Analysis, 8(2), 439-452.

Neal, R. M. (2011). MCMC using ensembles of states for problems with fast and slow variables such as Gaussian
process regression. arXiv:1101.0387.

Papaspiliopoulos, O., Roberts, G. O., & Sk&ouml;ld, M. (2007). A general framework for the parametrization of
hierarchical models. Statistical Science, 22(1), 59-73.

Tjelmeland, H. (2004). Using all Metropolis-Hastings proposals to estimate mean values. Statistics No. 4/2004,
Norwegian University of Science and Technology.

Yu, Y., & Meng, X.-L. (2011). To center or not to center: That is not the question -- an ancillarity-sufficiency
interweaving strategy (ASIS) for boosting MCMC efficiency. Journal of Computational and Graphical Statistics, 20(3),
531-570.
