---
title: "Sampler Validity"
description: "Why the particle step was biased in high dimensions, the exact and (in the sample stage) fixed kernel that replaces it, and the evidence that it is calibrated"
author: "Andrew Heathcote"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{"Sampler Validity"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
rm(list = ls())
library(EMC2)
set.seed(1)
```

This vignette is for whoever maintains the sampler. It documents a bias found on 2026-09-30 in the
particle-MCMC step (`new_particle()` in `R/sampling.R`), how to reproduce it in a few seconds on a
Gaussian toy and in a couple of minutes with an EMC2 fit, what was changed on the `dev-sampler`
branch, and the calibration evidence for the change. The old sampler is still in the package:

```{r, eval = FALSE}
options(emc.sampler = "legacy")   # the pre-2026-09-30 step, tuning, floors and fallbacks
options(emc.sampler = NULL)       # back to the current sampler
```

so every comparison below can be rerun with one switch.

## 1. Symptoms

Fitting a 29-35 parameter single-subject model to ~8000 trials (Ratcliff & Starns' recognition data
with the correlated multiple-threshold LNR, `vignette("rating-models")`) gave chains that never
left their start points (1-3% acceptance, Rhat up to 4), and where a fit did converge its posterior
SDs were 1.4-1.7x the Laplace SDs at the mode. A plain 25-parameter LNR showed the same, so this
was the sampler, not a likelihood. A grid of 48 cells (models with 9 / 25 / 29 / 35 parameters,
500 / 2000 / 8000 trials per subject, one subject with `type = "single"` and 8 subjects with
`type = "standard"`) then showed the pattern:

- 9 parameters: fine at every N, both types.
- 25-35 parameters, converged (Rhat 1.01): deviance excess `mean(D) - D(mode)` of 1.5-1.9 x k
  instead of k, posterior SD / Laplace SD 1.25-1.4 (max 1.67), for `single`; 1.3-1.6 x k and
  1.13-1.23 for `standard` at 8000 trials per subject.
- 29-35 parameters with >= 2000 trials: stuck (move rate .01, Rhat 1.2-2, posterior means up to
  18 posterior SDs from the truth).

The diagnostics used throughout are: **excess/k** = (mean sample-stage deviance - deviance at the
posterior mode) / number of parameters, which is 1 for a normal posterior; **SD ratio** = posterior
SD / Laplace SD (numerical Hessian of log-likelihood + log-prior at the mode); the **move rate**
(fraction of iterations whose value changed); Rhat; and z = (posterior mean - truth) / posterior SD
on simulated data.

## 2. The particle step as coded

Each iteration, for each subject, `new_particle()` draws one batch of particles from a mixture of
multivariate normals, keeps the chain's current value as particle 1, gives every particle the weight

    w(z) = exp( loglik(z) + logprior(z) - log q(z) )

with `q` the mixture density, and selects one particle in proportion to `w`. With `q` fixed this is
iterated sampling-importance-resampling (i-SIR, a conditional-SMC step with one time point) and it
leaves the posterior invariant. But the mixture is not fixed: in every stage some components are
centred on the current value `subj_mu` (preburn: both; burn: 2 of 3; adapt: 1 of 3; sample: 1 of 4,
the "random walk" `N(subj_mu, epsilon^2 chains_var)`), and `q` is evaluated at the current value
itself -- at the *peak* of those components. The current value's weight is therefore deflated by a
factor that depends on where the chain is, and the kernel is no longer posterior-invariant.

Two-particle sketch: with one candidate `z ~ N(x, S)` the move probability is
`pi(z) N(x|x,S) / (pi(z) N(x|x,S) + pi(x) N(z|x,S))`, and since `N(x|x,S) >= N(z|x,S)` this exceeds
the Barker probability `pi(z)/(pi(z) + pi(x))` by more the further `z` is from `x`. Detailed balance
fails, and the effect does not vanish with more particles when the importance weights are
degenerate -- which is exactly the high-dimensional, concentrated-posterior case.

## 3. Reproducing the bias on a Gaussian toy

The function below is the sample-stage step with the same particle numbers (`50 sqrt(d)`), the
same mixture weights (.05 prior, .3 random walk at the current value, .3 at a fixed mean, .35
"efficient" proposal) and a target that is the likelihood `N(0, I_d)` times a `N(0, 100 I_d)`
prior. `fix = TRUE` is the ensemble construction of section 4. Ten seconds per row.

```{r}
library(mvtnorm)
step <- function(x, d, eps, n, fix = FALSE) {
  z0 <- rep(0, d); I <- diag(d); P <- diag(100, d)
  # component 2 is the random walk: centred on x (as coded) or on an auxiliary centre c ~ N(x, S)
  S2 <- eps^2 * I
  c2 <- if (fix) drop(rmvnorm(1, x, S2)) else x
  Mus <- list(z0, c2, z0, z0); Ss <- list(P, S2, eps^2 * I, eps^2 * I); mix <- c(.05, .3, .3, .35)
  pn <- pmax(1, rmultinom(1, n, mix))
  props <- rbind(x, do.call(rbind, lapply(1:4, function(j) rmvnorm(pn[j], Mus[[j]], Ss[[j]]))))
  lpost <- dmvnorm(props, z0, I, log = TRUE) + dmvnorm(props, z0, P, log = TRUE)
  lq <- log(Reduce(`+`, lapply(1:4, function(j) mix[j] * dmvnorm(props, Mus[[j]], Ss[[j]]))))
  extra <- if (fix) dmvnorm(props, c2, S2, log = TRUE) else 0   # target factor N(z | c, S) of the ensemble move
  w <- exp(lpost + extra - lq - max(lpost + extra - lq))
  props[sample(nrow(props), 1, prob = w), ]
}
toy <- function(d, eps, fix = FALSE, iters = 1500, n = round(50 * sqrt(d))) {
  x <- rnorm(d); X <- matrix(NA, iters, d)
  for (i in 1:(300 + iters)) { x <- step(x, d, eps, n, fix); if (i > 300) X[i - 300, ] <- x }
  c(d = d, eps = eps, fix = fix, var_over_true = round(mean(apply(X, 2, var)) / (1 / (1 + 1/100)), 2))
}
rbind(toy(25, .6), toy(25, .3), toy(25, 2), toy(9, .6), toy(35, .6), toy(25, .6, fix = TRUE))
```

The chain "moves" every iteration but its variance is `eps^2`-ish, not the posterior's: narrower
components pull it down, wider ones push it up, more so as `d` grows. (In a longer run of this toy,
`rating-work/sampler/toy/`, the as-coded step gives 0.65 x the true variance at `eps = .6` -- the
floor the sample-stage adaptation pinned epsilon at -- 0.20 at .3, 1.47 at 2, and 2.1 for the burn
structure; the corrected step gives 0.98-1.02 everywhere.) At 9 parameters with well-scaled
proposals the bias is a few per cent, which is why simulation-based calibration on small models
passed and could not have caught it.

Three further things made real fits *stick* rather than just spread:

- `create_chain_proposals()` fell back to `chains_var = diag(.5)` when the covariance of the recent
  draws was not positive definite (a chain with few distinct draws) -- for a posterior with SD ~.02
  that is 35x too wide, so the chain cannot move, the next covariance is singular again, and the
  fallback persists.
- epsilon floors of .1 / .4 / .6 (preburn / burn-adapt / sample).
- The acceptance-driven epsilon update was also applied to the independence components
  (`chains_mu`, `eff`). For those, low acceptance means the proposal is too *narrow*, so shrinking
  it drove them to the floor; stuck fits ended with epsilon .6/.6/.6 and 94% of the mixture on an
  "efficient" proposal that accepted 0.05% of its particles.

## 4. What changed (`dev-sampler`)

**Kernel.** `preburn` and `burn`, whose draws are discarded, keep the old step: as a search it is a
fast stochastic hill-climb (it follows the proposal), and a valid random-walk kernel cannot travel
from prior-scale start points to a posterior with SD .02 in 35 dimensions in any reasonable time
(two exact-everywhere variants burned in 10-20x slower). `adapt` and `sample`, whose draws build the
proposals and the posterior, use an exact kernel: each iteration ONE of two kernels, chosen with the
mixture weights.

- *Local kernel* (the components centred on the current value): an ensemble move (Tjelmeland,
  2004; Neal, 2011). Draw an auxiliary centre `c ~ N(x, S)`, draw the particles from `N(c, S)`, and
  weight every particle -- the current value included -- by `pi(z) N(z | c, S) / q(z)` where `q` is
  the (now state-independent, given `c`) proposal density. The step targets the extended
  distribution `pi(x) N(c | x, S)`, whose `x`-marginal is `pi`, so it is exact.
- *Global kernel* (prior, `N(chains_mu, chains_var)`, `N(eff_mu, eff_var)`): the i-SIR step as
  before, valid because none of these depends on the current value.

Putting both in one batch with the ensemble correction is also exact, but the local factor
`N(z | c, S)` then penalises the global candidates; on the toy the mixture of kernels has 4-6x the
effective sample size per iteration (`rating-work/sampler/toy/isir_toy3.R`).

**The sample stage runs one fixed kernel (corrected 2026-10-01).** The first version of this
change (commit `002a5c6f`) stopped at the paragraph above, and an earlier version of this vignette
called its adapt/sample kernel exact. That was true of each iteration and not of the stage. The
sample stage kept re-choosing the kernel from the chain's own recent draws: `chains_var`,
`chains_mu`, `eff_mu` and `eff_var` were re-estimated at every step from the last <= 250
iterations, and epsilon and the mixing weights followed a window of acceptance counts. A chain
whose kernel depends on its recent past is not a Markov chain with the posterior as its stationary
distribution. Where the posterior's scale does not move under the adaptation the difference is
negligible, which is why the single-subject grid and the SBC of section 6 passed. In a
hierarchical funnel it is not. On a conjugate normal model with a known posterior
(25 parameters x 20 subjects; `rating-work/sampler/hier/stageH1/REPORT.md`), chains started at the
truth sank: for the weakly identified parameters the group SD had median 0.0036 against 0.159
exact, with 94% of the draws in the funnel's neck against 26%. With the same kernel held fixed
they did not sink.

The sample-stage kernel is therefore built and tuned at the end of adapt and then left alone:

- When adapt's stop criterion is met, the sample-stage proposals are made from the adapt draws and
  stored in `emc[[1]]$sample_kernel`. Every step of the sample stage, and every later call that
  adds samples, re-uses them.
- One more step of 100 iterations is run with that kernel, labelled `adapt`, in which its local
  epsilon, mixing weights and number of particles are adapted (`tune_sample_kernel()`). Epsilon is
  then set to its average over that step.
- In `sample` nothing adapts: `update_pm_settings()` is not called and `reset_pm_settings()`
  leaves the settings alone.

**Proposals that follow the group level (2026-10-01).** The same report found a second fault. The
local proposal was `N(c, epsilon^2 chains_var)` with one global scale. When the group variance of
some parameters collapses, the subjects' conditional posterior becomes 100x narrower in those
coordinates only, no single epsilon fits, and the step stops moving (move rate <= 0.001 with
epsilon on its floor of 0.01): 0 of 36 chain x parameter pairs left a collapsed start in 22000
iterations, against all 36 with an exact draw. The subject step conditions on the group level
`(mu, Sigma)`, so its proposals may depend on it. For every `type` with a group level they now do:

- `create_lik_prec()` estimates each subject's likelihood precision `L` (minus the Hessian of the
  log-likelihood) and the linear term of the same quadratic approximation, by central differences
  at the mean of the subject's recent draws (`lik_precision()`; `p^2 + p + 1` likelihood
  evaluations, with each parameter's step searched for so that the log-likelihood drops by about
  1). Moments of the draws cannot give this: when adapt runs in the neck the draws show the prior,
  not the likelihood.
- The local kernel's covariance is `epsilon^2 (L + Sigma^-1)^-1` at the current `Sigma`, in place
  of `epsilon^2 chains_var`. Epsilon is then a relative scale that no longer has to track the
  group level, and the 0.01 floor is not relied on.
- The global kernel's `N(chains_mu, chains_var)` component is replaced by the Gaussian
  approximation of the subject's conditional posterior at the current group level,
  `N((L + Sigma^-1)^-1 (l + Sigma^-1 mu), (L + Sigma^-1)^-1)` with `l` the linear term.
- The local kernel keeps at least a quarter of the sample-stage iterations.

`type = "single"` has no group level to follow and keeps `chains_var`.

**Tuning.** The local step's epsilon is adapted so that 3% of its particles beat the current value
(on the toy this is where ESS per iteration is maximal in 9, 25 and 35 dimensions; the old 25-30%
targets give steps 3-5x too small for the exact kernel); clamp [.01, 20] instead of the floors;
independence components are used at the scale of their estimated covariance (epsilon 1); burn's
`chains_var` epsilon, not its prior-variance one, is carried into adapt. In the exact kernels
epsilon is updated only in the iterations that used the local kernel, from that iteration's own
rate, with gain 0.1 and the update normalised by the target and clipped to [-1, 3]. (Updating it
every iteration from the rate accumulated over a window made it overshoot: over subjects it ended
anywhere between 0.01 and 5, which did not matter while it kept moving and does once it is
frozen.) preburn and burn keep the windowed counts, which restart every `step_size` iterations
(cumulative counts let epsilon run away). `update_epsilon_scale()` uses the square root of the
variance ratio (epsilon scales SDs). The `diag(.5)` fallback is replaced by the mean of the other
subjects' covariances, else the chain's previous one, else 0.1 x the prior variance diagonal.
`get_conditionals_single()` uses all (<= 750) adapt/sample draws rather than a random 250.

Everything above is switched by `options(emc.sampler = "legacy")`, which restores the old kernel in
every stage together with the old targets, floors, counts, fallback, 250-draw conditionals and the
sample-stage adaptation.

## 5. Reproducing it in EMC2

`vignettes/data/build_sampler_validity.R` simulates one subject (about 500 trials) from a
25-parameter LNR (`m, s ~ 0 + NWS:WF:lM, t0 ~ 1`, parameter values from a real participant), fits
it with both samplers using `fit()` defaults, and computes the Laplace SDs. Two minutes on three
cores; the results are cached.

```{r}
load("data/sampler-validity.RData")
tab <- t(sapply(summaries, function(s) c(
  `excess/k` = round(s$excess / s$k, 2),
  `SD/Laplace (median)` = round(median(s$pars$post_sd / s$pars$laplace_sd), 2),
  `SD/Laplace (min)` = round(min(s$pars$post_sd / s$pars$laplace_sd), 2),
  `SD/Laplace (max)` = round(max(s$pars$post_sd / s$pars$laplace_sd), 2),
  `max |z|` = round(max(abs(s$pars$z)), 1), `move rate` = round(s$move_rate, 2),
  `max Rhat` = round(s$max_rhat, 3), iterations = sum(s$iters))))
tab
```

```{r, fig.width = 7, fig.height = 4}
r <- sapply(summaries, function(s) s$pars$post_sd / s$pars$laplace_sd)
plot(r[, "legacy"], type = "h", lwd = 6, col = "grey60", ylim = c(0, max(r) * 1.05), xaxt = "n",
     xlab = "parameter", ylab = "posterior SD / Laplace SD", main = "25-parameter LNR, ~500 trials")
points(seq_len(nrow(r)) + .25, r[, "current"], type = "h", lwd = 6, col = "#0072B2")
abline(h = 1, lty = 2); axis(1, at = seq_len(nrow(r)), labels = FALSE)
legend("topleft", c("legacy sampler", "current sampler"), fill = c("grey60", "#0072B2"), bty = "n")
```

The legacy sampler reports Rhat ~1 and a move rate of 1.00 and is wrong by a third in every SD; the
current sampler's honest move rate is about .8. The per-stage tuning state (`epsilon`, `mix`,
`n_particles`) is in `attr(emc[[1]]$samples, "pm_settings")`:

```{r}
sapply(summaries, function(s) round(s$epsilon, 2))
```

With the legacy sampler the independence components end at an epsilon above 1 (1.2-1.6 from run
to run; wider than the posterior, which is where the over-dispersion comes from: the chain follows
them); with the current one they are fixed at 1 and only the local step is scaled (about
.4-.5 x `chains_var`).

To go further -- more parameters, more trials, `type = "standard"` -- change the design and
`n_trials` in the build script; the full grid, its runner and per-stage snapshots of the tuning
state are in `rating-work/sampler/grid/` (outside the package).

## 6. Evidence that the current sampler is calibrated

**Grid** (`rating-work/sampler/REPORT.md`, sections 1 and 5; 3 chains, `fit()` defaults, 1000 sample
iterations). Single subject, excess/k and median SD/Laplace, legacy -> current:

| k | N = 500 | N = 2000 | N = 8000 |
|---|---|---|---|
| 9 | 1.00 / 1.04 -> 0.90 / 0.99 | 1.09 / 1.06 -> 0.95 / 0.99 | 1.09 / 1.06 -> 0.99 / 1.00 |
| 25 | 1.68 / 1.34 -> 0.94 / 1.01 | 1.51 / 1.25 -> 0.97 / 0.99 | stuck -> 1.00 / 1.00 |
| 29 | 1.92 / 1.39 -> 0.92 / 0.96 | stuck -> 0.98 / 0.97 | stuck -> 0.99 / 1.00 |
| 35 | 1.92 / 1.38 -> 0.92 / 0.99 | stuck -> 0.92 / 0.92 | stuck -> 0.95 / 0.98 |

("stuck" = burn never converged, sample-stage move rate .01.) Rhat <= 1.10 in every current-sampler
cell, burn-in 100-300 iterations as before, and the 35-parameter x 8000-trial fit takes 22 minutes.
Eight subjects with `type = "standard"`: every cell with >= 2000 trials per subject is calibrated
(excess/k 0.91-1.02, SD ratio 0.89-0.99; legacy 1.27-1.64 and 1.09-1.23 at k >= 25). The
8-subject x 500-trial cells with 25-35 parameters do not reach Rhat < 1.1 under `fit()` defaults
with either sampler (see section 7).

**Simulation-based calibration.** The MTLNR SBC (8 single-subject cells x 500 replicates, one
15-subject cell x 100; `SBC-archive/2026-09-30_dev-sampler_MTLNR-sbc-regression`) rerun with the
current sampler: no KS p below .01, Fisher's combined p over the 68 parameter x cell tests .15 (the
legacy run: .051), 95% coverage .92-.98 centred on nominal (legacy ran high, .93-.98), |std_bias|
<= .17, hierarchical mu / Sigma ranks uniform; 3.5 h instead of 5.8 h.

**Real data.** The five Ratcliff-Starns models of `vignette("rating-models")`'s source paper
(24-35 parameters x 11 subjects, ~7600 trials each) all converge in the standard 1000 sample
iterations; EMC2's DIC ordering equals the paper's, EMC2 finds lower deviance minima than the
paper's DE-MCMC fits did, and the parameter estimates correlate >= .98 with them.

**After the 2026-10-01 changes of section 4** (fixed sample-stage kernel, proposals that follow
the group level; `rating-work/sampler/hier/stageH1b/REPORT.md`):

- *Conjugate model with a known posterior* (25 parameters x 20 subjects, 12 of them weakly
  identified, 20000 sample iterations x 6 chains from the truth and 6 from a collapsed start).
  The posterior of the weakly identified group SDs is reproduced from both starts: medians
  0.146-0.170 against 0.159 exact, share of draws in the funnel's neck 0.18-0.27 against 0.18-0.26,
  subject-level posterior SDs 0.97-1.01 x exact. With the group level held fixed the subject step
  alone is exact at the truth and in the collapsed state (SD ratio 1.00, about 2 iterations per
  effective draw; before the change it did not move in the collapsed state).
- *The grid's single-subject cells* (9, 25 and 35 parameters x 500 / 2000 / 8000 trials), rerun:
  excess/k 0.90-1.00, median SD/Laplace 0.98-1.03, Rhat <= 1.09. The 25-parameter cells take
  1.3-1.5x as long, because the number of particles is now set once, at the end of adapt.
- *A hierarchical cell with no funnel* (25 parameters x 20 subjects x 2000 trials): Rhat 1.02,
  excess/k 1.03, as before.

**Tests.** The suite passes (1434); the three snapshot tests in `test-fit.R` that pin specific
draws were re-recorded, as any change to the sampler requires.

## 7. What is still open

An earlier version of this section said that the hierarchical fits which do not converge
(few trials per weakly identified parameter: 8-40 subjects x 200-500 trials x 25-35 parameters,
alpha Rhat 1.4-1.9 under `fit()` defaults) were stuck in a funnel because of the group-level Gibbs
step, not the particle kernel. That was wrong. On a conjugate model with a known posterior
(`rating-work/sampler/hier/stageH1/REPORT.md`) EMC2's group step with an exact draw of the subject
parameters reproduces the exact posterior, and the failure was the particle step: the sample
stage's continued adaptation made chains sink into the funnel's neck from healthy states, and
the local proposal froze once they were there. Both are repaired (section 4). On that model
`fit()` defaults used to stop at max tries with the weak parameters' group SDs at a third to two
thirds of the exact median. They still stop at max tries (Rhat 1.3-2.4 after 1700-1900 sample
iterations), but the draws now agree with the exact posterior (medians 0.15-0.18 against 0.16,
neck shares 0.16-0.24 against 0.18-0.26): there are just few effective ones, as there are with an
exact draw of the subject parameters (Rhat 1.3 under the same protocol).

What remains is slow, not wrong:

- The centred alternation "subjects given the group level, group level given the subjects" mixes
  slowly in a funnel even with a perfect subject step: on the conjugate model one effective draw
  of a weak parameter's log group SD takes about 500 iterations, and leaving a collapsed state
  takes about 400. The repaired particle step takes about 1.3x as long as an exact draw to leave
  it. So `fit()`'s 1000 sample iterations give few effective draws of those group SDs, and Rhat
  for them can stay above 1.1 (the 25-parameter x 20-subject x 200-trial LNR cell stops at alpha
  Rhat 1.16 and group-SD Rhat 1.5, where it was 1.64 and 2.15). An interweaving (non-centred)
  scale move is the planned remedy.
- `init()` starts every chain in the neck (subjects bunched at the group mean, group SD about
  0.14 for every parameter), and `preburn`/`burn` keep the old search step.
- Only `type = "standard"` has been checked against an exact reference. The other types with a
  group level share the code path.

## References

Neal, R. M. (2011). MCMC using ensembles of states for problems with fast and slow variables such
as Gaussian process regression. arXiv:1101.0387.

Tjelmeland, H. (2004). Using all Metropolis-Hastings proposals to estimate mean values. Statistics
No. 4/2004, Norwegian University of Science and Technology.
