---
title: "Sampler Validity"
description: "Why the particle step was biased in high dimensions, the exact kernel that replaces it in adapt/sample, and the evidence that it is calibrated"
author: "Andrew Heathcote"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{"Sampler Validity"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
rm(list = ls())
library(EMC2)
set.seed(1)
```

This vignette is for whoever maintains the sampler. It documents a bias found on 2026-09-30 in the
particle-MCMC step (`new_particle()` in `R/sampling.R`), how to reproduce it in a few seconds on a
Gaussian toy and in a couple of minutes with an EMC2 fit, what was changed on the `dev-sampler`
branch, and the calibration evidence for the change. The old sampler is still in the package:

```{r, eval = FALSE}
options(emc.sampler = "legacy")   # the pre-2026-09-30 step, tuning, floors and fallbacks
options(emc.sampler = NULL)       # back to the current sampler
```

so every comparison below can be rerun with one switch.

## 1. Symptoms

Fitting a 29-35 parameter single-subject model to ~8000 trials (Ratcliff & Starns' recognition data
with the correlated multiple-threshold LNR, `vignette("rating-models")`) gave chains that never
left their start points (1-3% acceptance, Rhat up to 4), and where a fit did converge its posterior
SDs were 1.4-1.7x the Laplace SDs at the mode. A plain 25-parameter LNR showed the same, so this
was the sampler, not a likelihood. A grid of 48 cells (models with 9 / 25 / 29 / 35 parameters,
500 / 2000 / 8000 trials per subject, one subject with `type = "single"` and 8 subjects with
`type = "standard"`) then showed the pattern:

- 9 parameters: fine at every N, both types.
- 25-35 parameters, converged (Rhat 1.01): deviance excess `mean(D) - D(mode)` of 1.5-1.9 x k
  instead of k, posterior SD / Laplace SD 1.25-1.4 (max 1.67), for `single`; 1.3-1.6 x k and
  1.13-1.23 for `standard` at 8000 trials per subject.
- 29-35 parameters with >= 2000 trials: stuck (move rate .01, Rhat 1.2-2, posterior means up to
  18 posterior SDs from the truth).

The diagnostics used throughout are: **excess/k** = (mean sample-stage deviance - deviance at the
posterior mode) / number of parameters, which is 1 for a normal posterior; **SD ratio** = posterior
SD / Laplace SD (numerical Hessian of log-likelihood + log-prior at the mode); the **move rate**
(fraction of iterations whose value changed); Rhat; and z = (posterior mean - truth) / posterior SD
on simulated data.

## 2. The particle step as coded

Each iteration, for each subject, `new_particle()` draws one batch of particles from a mixture of
multivariate normals, keeps the chain's current value as particle 1, gives every particle the weight

    w(z) = exp( loglik(z) + logprior(z) - log q(z) )

with `q` the mixture density, and selects one particle in proportion to `w`. With `q` fixed this is
iterated sampling-importance-resampling (i-SIR, a conditional-SMC step with one time point) and it
leaves the posterior invariant. But the mixture is not fixed: in every stage some components are
centred on the current value `subj_mu` (preburn: both; burn: 2 of 3; adapt: 1 of 3; sample: 1 of 4,
the "random walk" `N(subj_mu, epsilon^2 chains_var)`), and `q` is evaluated at the current value
itself -- at the *peak* of those components. The current value's weight is therefore deflated by a
factor that depends on where the chain is, and the kernel is no longer posterior-invariant.

Two-particle sketch: with one candidate `z ~ N(x, S)` the move probability is
`pi(z) N(x|x,S) / (pi(z) N(x|x,S) + pi(x) N(z|x,S))`, and since `N(x|x,S) >= N(z|x,S)` this exceeds
the Barker probability `pi(z)/(pi(z) + pi(x))` by more the further `z` is from `x`. Detailed balance
fails, and the effect does not vanish with more particles when the importance weights are
degenerate -- which is exactly the high-dimensional, concentrated-posterior case.

## 3. Reproducing the bias on a Gaussian toy

The function below is the sample-stage step with the same particle numbers (`50 sqrt(d)`), the
same mixture weights (.05 prior, .3 random walk at the current value, .3 at a fixed mean, .35
"efficient" proposal) and a target that is the likelihood `N(0, I_d)` times a `N(0, 100 I_d)`
prior. `fix = TRUE` is the ensemble construction of section 4. Ten seconds per row.

```{r}
library(mvtnorm)
step <- function(x, d, eps, n, fix = FALSE) {
  z0 <- rep(0, d); I <- diag(d); P <- diag(100, d)
  # component 2 is the random walk: centred on x (as coded) or on an auxiliary centre c ~ N(x, S)
  S2 <- eps^2 * I
  c2 <- if (fix) drop(rmvnorm(1, x, S2)) else x
  Mus <- list(z0, c2, z0, z0); Ss <- list(P, S2, eps^2 * I, eps^2 * I); mix <- c(.05, .3, .3, .35)
  pn <- pmax(1, rmultinom(1, n, mix))
  props <- rbind(x, do.call(rbind, lapply(1:4, function(j) rmvnorm(pn[j], Mus[[j]], Ss[[j]]))))
  lpost <- dmvnorm(props, z0, I, log = TRUE) + dmvnorm(props, z0, P, log = TRUE)
  lq <- log(Reduce(`+`, lapply(1:4, function(j) mix[j] * dmvnorm(props, Mus[[j]], Ss[[j]]))))
  extra <- if (fix) dmvnorm(props, c2, S2, log = TRUE) else 0   # target factor N(z | c, S) of the ensemble move
  w <- exp(lpost + extra - lq - max(lpost + extra - lq))
  props[sample(nrow(props), 1, prob = w), ]
}
toy <- function(d, eps, fix = FALSE, iters = 1500, n = round(50 * sqrt(d))) {
  x <- rnorm(d); X <- matrix(NA, iters, d)
  for (i in 1:(300 + iters)) { x <- step(x, d, eps, n, fix); if (i > 300) X[i - 300, ] <- x }
  c(d = d, eps = eps, fix = fix, var_over_true = round(mean(apply(X, 2, var)) / (1 / (1 + 1/100)), 2))
}
rbind(toy(25, .6), toy(25, .3), toy(25, 2), toy(9, .6), toy(35, .6), toy(25, .6, fix = TRUE))
```

The chain "moves" every iteration but its variance is `eps^2`-ish, not the posterior's: narrower
components pull it down, wider ones push it up, more so as `d` grows. (In a longer run of this toy,
`rating-work/sampler/toy/`, the as-coded step gives 0.65 x the true variance at `eps = .6` -- the
floor the sample-stage adaptation pinned epsilon at -- 0.20 at .3, 1.47 at 2, and 2.1 for the burn
structure; the corrected step gives 0.98-1.02 everywhere.) At 9 parameters with well-scaled
proposals the bias is a few per cent, which is why simulation-based calibration on small models
passed and could not have caught it.

Three further things made real fits *stick* rather than just spread:

- `create_chain_proposals()` fell back to `chains_var = diag(.5)` when the covariance of the recent
  draws was not positive definite (a chain with few distinct draws) -- for a posterior with SD ~.02
  that is 35x too wide, so the chain cannot move, the next covariance is singular again, and the
  fallback persists.
- epsilon floors of .1 / .4 / .6 (preburn / burn-adapt / sample).
- The acceptance-driven epsilon update was also applied to the independence components
  (`chains_mu`, `eff`). For those, low acceptance means the proposal is too *narrow*, so shrinking
  it drove them to the floor; stuck fits ended with epsilon .6/.6/.6 and 94% of the mixture on an
  "efficient" proposal that accepted 0.05% of its particles.

## 4. What changed (`dev-sampler`)

**Kernel.** `preburn` and `burn`, whose draws are discarded, keep the old step: as a search it is a
fast stochastic hill-climb (it follows the proposal), and a valid random-walk kernel cannot travel
from prior-scale start points to a posterior with SD .02 in 35 dimensions in any reasonable time
(two exact-everywhere variants burned in 10-20x slower). `adapt` and `sample`, whose draws build the
proposals and the posterior, use an exact kernel: each iteration ONE of two kernels, chosen with the
mixture weights.

- *Local kernel* (the components centred on the current value): an ensemble move (Tjelmeland,
  2004; Neal, 2011). Draw an auxiliary centre `c ~ N(x, S)`, draw the particles from `N(c, S)`, and
  weight every particle -- the current value included -- by `pi(z) N(z | c, S) / q(z)` where `q` is
  the (now state-independent, given `c`) proposal density. The step targets the extended
  distribution `pi(x) N(c | x, S)`, whose `x`-marginal is `pi`, so it is exact.
- *Global kernel* (prior, `N(chains_mu, chains_var)`, `N(eff_mu, eff_var)`): the i-SIR step as
  before, valid because none of these depends on the current value.

Putting both in one batch with the ensemble correction is also exact, but the local factor
`N(z | c, S)` then penalises the global candidates; on the toy the mixture of kernels has 4-6x the
effective sample size per iteration (`rating-work/sampler/toy/isir_toy3.R`).

**Tuning.** The local step's epsilon is adapted so that 3% of its particles beat the current value
(on the toy this is where ESS per iteration is maximal in 9, 25 and 35 dimensions; the old 25-30%
targets give steps 3-5x too small for the exact kernel); the update step is normalised by the
target and clipped to [-1, 1]; clamp [.01, 20] instead of the floors; acceptance counts restart
every `step_size` iterations in every stage (cumulative counts let epsilon run away); independence
components are used at the scale of their estimated covariance (epsilon 1); burn's `chains_var`
epsilon, not its prior-variance one, is carried into adapt. `update_epsilon_scale()` uses the square
root of the variance ratio (epsilon scales SDs). The `diag(.5)` fallback is replaced by the mean of
the other subjects' covariances, else the chain's previous one, else 0.1 x the prior variance
diagonal. `get_conditionals_single()` uses all (<= 750) adapt/sample draws rather than a random 250.

Everything above is switched by `options(emc.sampler = "legacy")`, which restores the old kernel in
every stage together with the old targets, floors, counts, fallback and 250-draw conditionals.

## 5. Reproducing it in EMC2

`vignettes/data/build_sampler_validity.R` simulates one subject (about 500 trials) from a
25-parameter LNR (`m, s ~ 0 + NWS:WF:lM, t0 ~ 1`, parameter values from a real participant), fits
it with both samplers using `fit()` defaults, and computes the Laplace SDs. Two minutes on three
cores; the results are cached.

```{r}
load("data/sampler-validity.RData")
tab <- t(sapply(summaries, function(s) c(
  `excess/k` = round(s$excess / s$k, 2),
  `SD/Laplace (median)` = round(median(s$pars$post_sd / s$pars$laplace_sd), 2),
  `SD/Laplace (min)` = round(min(s$pars$post_sd / s$pars$laplace_sd), 2),
  `SD/Laplace (max)` = round(max(s$pars$post_sd / s$pars$laplace_sd), 2),
  `max |z|` = round(max(abs(s$pars$z)), 1), `move rate` = round(s$move_rate, 2),
  `max Rhat` = round(s$max_rhat, 3), iterations = sum(s$iters))))
tab
```

```{r, fig.width = 7, fig.height = 4}
r <- sapply(summaries, function(s) s$pars$post_sd / s$pars$laplace_sd)
plot(r[, "legacy"], type = "h", lwd = 6, col = "grey60", ylim = c(0, max(r) * 1.05), xaxt = "n",
     xlab = "parameter", ylab = "posterior SD / Laplace SD", main = "25-parameter LNR, ~500 trials")
points(seq_len(nrow(r)) + .25, r[, "current"], type = "h", lwd = 6, col = "#0072B2")
abline(h = 1, lty = 2); axis(1, at = seq_len(nrow(r)), labels = FALSE)
legend("topleft", c("legacy sampler", "current sampler"), fill = c("grey60", "#0072B2"), bty = "n")
```

The legacy sampler reports Rhat ~1 and a move rate of 1.00 and is wrong by a third in every SD; the
current sampler's honest move rate is about .8. The per-stage tuning state (`epsilon`, `mix`,
`n_particles`) is in `attr(emc[[1]]$samples, "pm_settings")`:

```{r}
sapply(summaries, function(s) round(s$epsilon, 2))
```

With the legacy sampler the independence components end at epsilon 1.5-1.6 (wider than the
posterior, which is where the over-dispersion comes from: the chain follows them); with the current
one they are fixed at 1 and only the local step is scaled (about .5 x `chains_var`).

To go further -- more parameters, more trials, `type = "standard"` -- change the design and
`n_trials` in the build script; the full grid, its runner and per-stage snapshots of the tuning
state are in `rating-work/sampler/grid/` (outside the package).

## 6. Evidence that the current sampler is calibrated

**Grid** (`rating-work/sampler/REPORT.md`, sections 1 and 5; 3 chains, `fit()` defaults, 1000 sample
iterations). Single subject, excess/k and median SD/Laplace, legacy -> current:

| k | N = 500 | N = 2000 | N = 8000 |
|---|---|---|---|
| 9 | 1.00 / 1.04 -> 0.90 / 0.99 | 1.09 / 1.06 -> 0.95 / 0.99 | 1.09 / 1.06 -> 0.99 / 1.00 |
| 25 | 1.68 / 1.34 -> 0.94 / 1.01 | 1.51 / 1.25 -> 0.97 / 0.99 | stuck -> 1.00 / 1.00 |
| 29 | 1.92 / 1.39 -> 0.92 / 0.96 | stuck -> 0.98 / 0.97 | stuck -> 0.99 / 1.00 |
| 35 | 1.92 / 1.38 -> 0.92 / 0.99 | stuck -> 0.92 / 0.92 | stuck -> 0.95 / 0.98 |

("stuck" = burn never converged, sample-stage move rate .01.) Rhat <= 1.10 in every current-sampler
cell, burn-in 100-300 iterations as before, and the 35-parameter x 8000-trial fit takes 22 minutes.
Eight subjects with `type = "standard"`: every cell with >= 2000 trials per subject is calibrated
(excess/k 0.91-1.02, SD ratio 0.89-0.99; legacy 1.27-1.64 and 1.09-1.23 at k >= 25). The
8-subject x 500-trial cells with 25-35 parameters do not reach Rhat < 1.1 under `fit()` defaults
with either sampler (see section 7).

**Simulation-based calibration.** The MTLNR SBC (8 single-subject cells x 500 replicates, one
15-subject cell x 100; `SBC-archive/2026-09-30_dev-sampler_MTLNR-sbc-regression`) rerun with the
current sampler: no KS p below .01, Fisher's combined p over the 68 parameter x cell tests .15 (the
legacy run: .051), 95% coverage .92-.98 centred on nominal (legacy ran high, .93-.98), |std_bias|
<= .17, hierarchical mu / Sigma ranks uniform; 3.5 h instead of 5.8 h.

**Real data.** The five Ratcliff-Starns models of `vignette("rating-models")`'s source paper
(24-35 parameters x 11 subjects, ~7600 trials each) all converge in the standard 1000 sample
iterations; EMC2's DIC ordering equals the paper's, EMC2 finds lower deviance minima than the
paper's DE-MCMC fits did, and the parameter estimates correlate >= .98 with them.

**Tests.** The suite passes (1409); the three snapshot tests in `test-fit.R` that pin specific
draws were re-recorded, as any change to the sampler requires.

## 7. What is still open

With 8 subjects, 500 trials each and 25-35 parameters the hierarchical fits do not converge under
`fit()` defaults (alpha Rhat 1.4-1.9), and a 4000-iteration run still sits at 1.2. The chains are in
a "funnel": the group variance of loser-accumulator means that each subject's data barely identify
(~8 losing trials) collapses, all subjects are pinned at the group mean, and the Gibbs step for the
group level cannot get out. The legacy sampler reports Rhat 1.1 in two of these cells because its
over-dispersed subject draws keep the group variance from collapsing -- a self-consistent but wrong
equilibrium. This is a property of the group-level Gibbs step, not of the particle kernel, and is
being worked on separately.

## References

Neal, R. M. (2011). MCMC using ensembles of states for problems with fast and slow variables such
as Gaussian process regression. arXiv:1101.0387.

Tjelmeland, H. (2004). Using all Metropolis-Hastings proposals to estimate mean values. Statistics
No. 4/2004, Norwegian University of Science and Technology.
