---
title: "Sampler Validity"
description: "Why the particle step was biased in high dimensions, the exact and (in the sample stage) fixed kernel that replaces it, and the evidence that it is calibrated"
author: "Andrew Heathcote"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{"Sampler Validity"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
rm(list = ls())
library(EMC2)
set.seed(1)
```

This vignette is for whoever maintains the sampler. It documents a bias found on 2026-09-30 in the
particle-MCMC step (`new_particle()` in `R/sampling.R`), how to reproduce it in a few seconds on a
Gaussian toy and in a couple of minutes with an EMC2 fit, what was changed on the `dev-sampler`
branch, and the calibration evidence for the change. The old sampler is still in the package:

```{r, eval = FALSE}
options(emc.sampler = "legacy")   # the pre-2026-09-30 step, tuning, floors and fallbacks
options(emc.sampler = NULL)       # back to the current sampler
```

so every comparison below can be rerun with one switch.

## 1. Symptoms

Fitting a 29-35 parameter single-subject model to ~8000 trials (Ratcliff & Starns' recognition data
with the correlated multiple-threshold LNR, `vignette("rating-models")`) gave chains that never
left their start points (1-3% acceptance, Rhat up to 4), and where a fit did converge its posterior
SDs were 1.4-1.7x the Laplace SDs at the mode. A plain 25-parameter LNR showed the same, so this
was the sampler, not a likelihood. A grid of 48 cells (models with 9 / 25 / 29 / 35 parameters,
500 / 2000 / 8000 trials per subject, one subject with `type = "single"` and 8 subjects with
`type = "standard"`) then showed the pattern:

- 9 parameters: fine at every N, both types.
- 25-35 parameters, converged (Rhat 1.01): deviance excess `mean(D) - D(mode)` of 1.5-1.9 x k
  instead of k, posterior SD / Laplace SD 1.25-1.4 (max 1.67), for `single`; 1.3-1.6 x k and
  1.13-1.23 for `standard` at 8000 trials per subject.
- 29-35 parameters with >= 2000 trials: stuck (move rate .01, Rhat 1.2-2, posterior means up to
  18 posterior SDs from the truth).

The diagnostics used throughout are: **excess/k** = (mean sample-stage deviance - deviance at the
posterior mode) / number of parameters, which is 1 for a normal posterior; **SD ratio** = posterior
SD / Laplace SD (numerical Hessian of log-likelihood + log-prior at the mode); the **move rate**
(fraction of iterations whose value changed); Rhat; and z = (posterior mean - truth) / posterior SD
on simulated data.

## 2. The particle step as coded

Each iteration, for each subject, `new_particle()` draws one batch of particles from a mixture of
multivariate normals, keeps the chain's current value as particle 1, gives every particle the weight

    w(z) = exp( loglik(z) + logprior(z) - log q(z) )

with `q` the mixture density, and selects one particle in proportion to `w`. With `q` fixed this is
iterated sampling-importance-resampling (i-SIR, a conditional-SMC step with one time point) and it
leaves the posterior invariant. But the mixture is not fixed: in every stage some components are
centred on the current value `subj_mu` (preburn: both; burn: 2 of 3; adapt: 1 of 3; sample: 1 of 4,
the "random walk" `N(subj_mu, epsilon^2 chains_var)`), and `q` is evaluated at the current value
itself -- at the *peak* of those components. The current value's weight is therefore deflated by a
factor that depends on where the chain is, and the kernel is no longer posterior-invariant.

Two-particle sketch: with one candidate `z ~ N(x, S)` the move probability is
`pi(z) N(x|x,S) / (pi(z) N(x|x,S) + pi(x) N(z|x,S))`, and since `N(x|x,S) >= N(z|x,S)` this exceeds
the Barker probability `pi(z)/(pi(z) + pi(x))` by more the further `z` is from `x`. Detailed balance
fails, and the effect does not vanish with more particles when the importance weights are
degenerate -- which is exactly the high-dimensional, concentrated-posterior case.

## 3. Reproducing the bias on a Gaussian toy

The function below is the sample-stage step with the same particle numbers (`50 sqrt(d)`), the
same mixture weights (.05 prior, .3 random walk at the current value, .3 at a fixed mean, .35
"efficient" proposal) and a target that is the likelihood `N(0, I_d)` times a `N(0, 100 I_d)`
prior. `fix = TRUE` is the ensemble construction of section 4. Ten seconds per row.

```{r}
library(mvtnorm)
step <- function(x, d, eps, n, fix = FALSE) {
  z0 <- rep(0, d); I <- diag(d); P <- diag(100, d)
  # component 2 is the random walk: centred on x (as coded) or on an auxiliary centre c ~ N(x, S)
  S2 <- eps^2 * I
  c2 <- if (fix) drop(rmvnorm(1, x, S2)) else x
  Mus <- list(z0, c2, z0, z0); Ss <- list(P, S2, eps^2 * I, eps^2 * I); mix <- c(.05, .3, .3, .35)
  pn <- pmax(1, rmultinom(1, n, mix))
  props <- rbind(x, do.call(rbind, lapply(1:4, function(j) rmvnorm(pn[j], Mus[[j]], Ss[[j]]))))
  lpost <- dmvnorm(props, z0, I, log = TRUE) + dmvnorm(props, z0, P, log = TRUE)
  lq <- log(Reduce(`+`, lapply(1:4, function(j) mix[j] * dmvnorm(props, Mus[[j]], Ss[[j]]))))
  extra <- if (fix) dmvnorm(props, c2, S2, log = TRUE) else 0   # target factor N(z | c, S) of the ensemble move
  w <- exp(lpost + extra - lq - max(lpost + extra - lq))
  props[sample(nrow(props), 1, prob = w), ]
}
toy <- function(d, eps, fix = FALSE, iters = 1500, n = round(50 * sqrt(d))) {
  x <- rnorm(d); X <- matrix(NA, iters, d)
  for (i in 1:(300 + iters)) { x <- step(x, d, eps, n, fix); if (i > 300) X[i - 300, ] <- x }
  c(d = d, eps = eps, fix = fix, var_over_true = round(mean(apply(X, 2, var)) / (1 / (1 + 1/100)), 2))
}
rbind(toy(25, .6), toy(25, .3), toy(25, 2), toy(9, .6), toy(35, .6), toy(25, .6, fix = TRUE))
```

The chain "moves" every iteration but its variance is `eps^2`-ish, not the posterior's: narrower
components pull it down, wider ones push it up, more so as `d` grows. (In a longer run of this toy,
`rating-work/sampler/toy/`, the as-coded step gives 0.65 x the true variance at `eps = .6` -- the
floor the sample-stage adaptation pinned epsilon at -- 0.20 at .3, 1.47 at 2, and 2.1 for the burn
structure; the corrected step gives 0.98-1.02 everywhere.) At 9 parameters with well-scaled
proposals the bias is a few per cent, which is why simulation-based calibration on small models
passed and could not have caught it.

Three further things made real fits *stick* rather than just spread:

- `create_chain_proposals()` fell back to `chains_var = diag(.5)` when the covariance of the recent
  draws was not positive definite (a chain with few distinct draws) -- for a posterior with SD ~.02
  that is 35x too wide, so the chain cannot move, the next covariance is singular again, and the
  fallback persists.
- epsilon floors of .1 / .4 / .6 (preburn / burn-adapt / sample).
- The acceptance-driven epsilon update was also applied to the independence components
  (`chains_mu`, `eff`). For those, low acceptance means the proposal is too *narrow*, so shrinking
  it drove them to the floor; stuck fits ended with epsilon .6/.6/.6 and 94% of the mixture on an
  "efficient" proposal that accepted 0.05% of its particles.

## 4. What changed (`dev-sampler`)

**Kernel.** `preburn` and `burn`, whose draws are discarded, keep the old step: as a search it is a
fast stochastic hill-climb (it follows the proposal), and a valid random-walk kernel cannot travel
from prior-scale start points to a posterior with SD .02 in 35 dimensions in any reasonable time
(two exact-everywhere variants burned in 10-20x slower). `adapt` and `sample`, whose draws build the
proposals and the posterior, use an exact kernel: each iteration ONE of two kernels, chosen with the
mixture weights.

- *Local kernel* (the components centred on the current value): an ensemble move (Tjelmeland,
  2004; Neal, 2011). Draw an auxiliary centre `c ~ N(x, S)`, draw the particles from `N(c, S)`, and
  weight every particle -- the current value included -- by `pi(z) N(z | c, S) / q(z)` where `q` is
  the (now state-independent, given `c`) proposal density. The step targets the extended
  distribution `pi(x) N(c | x, S)`, whose `x`-marginal is `pi`, so it is exact.
- *Global kernel* (prior, `N(chains_mu, chains_var)`, `N(eff_mu, eff_var)`): the i-SIR step as
  before, valid because none of these depends on the current value.

Putting both in one batch with the ensemble correction is also exact, but the local factor
`N(z | c, S)` then penalises the global candidates; on the toy the mixture of kernels has 4-6x the
effective sample size per iteration (`rating-work/sampler/toy/isir_toy3.R`).

**The sample stage runs one fixed kernel (corrected 2026-10-01).** The first version of this
change (commit `002a5c6f`) stopped at the paragraph above, and an earlier version of this vignette
called its adapt/sample kernel exact. That was true of each iteration and not of the stage. The
sample stage kept re-choosing the kernel from the chain's own recent draws: `chains_var`,
`chains_mu`, `eff_mu` and `eff_var` were re-estimated at every step from the last <= 250
iterations, and epsilon and the mixing weights followed a window of acceptance counts. A chain
whose kernel depends on its recent past is not a Markov chain with the posterior as its stationary
distribution. Where the posterior's scale does not move under the adaptation the difference is
negligible, which is why the single-subject grid and the SBC of section 6 passed. In a
hierarchical funnel it is not. On a conjugate normal model with a known posterior
(25 parameters x 20 subjects; `rating-work/sampler/hier/stageH1/REPORT.md`), chains started at the
truth sank: for the weakly identified parameters the group SD had median 0.0036 against 0.159
exact, with 94% of the draws in the funnel's neck against 26%. With the same kernel held fixed
they did not sink.

The sample-stage kernel is therefore built and tuned at the end of adapt and then left alone:

- When adapt's stop criterion is met, the sample-stage proposals are made from the adapt draws and
  stored in `emc[[1]]$sample_kernel`. Every step of the sample stage, and every later call that
  adds samples, re-uses them.
- One more step of 100 iterations is run with that kernel, labelled `adapt`, in which its local
  epsilon, mixing weights and number of particles are adapted (`tune_sample_kernel()`). Epsilon is
  then set to its average over that step.
- In `sample` nothing adapts: `update_pm_settings()` is not called and `reset_pm_settings()`
  leaves the settings alone.

**Proposals that follow the group level (2026-10-01).** The same report found a second fault. The
local proposal was `N(c, epsilon^2 chains_var)` with one global scale. When the group variance of
some parameters collapses, the subjects' conditional posterior becomes 100x narrower in those
coordinates only, no single epsilon fits, and the step stops moving (move rate <= 0.001 with
epsilon on its floor of 0.01): 0 of 36 chain x parameter pairs left a collapsed start in 22000
iterations, against all 36 with an exact draw. The subject step conditions on the group level
`(mu, Sigma)`, so its proposals may depend on it. For every `type` with a group level they now do:

- `create_lik_prec()` estimates each subject's likelihood precision `L` (minus the Hessian of the
  log-likelihood) and the linear term of the same quadratic approximation, by central differences
  at the mean of the subject's recent draws (`lik_precision()`; `p^2 + p + 1` likelihood
  evaluations, with each parameter's step searched for so that the log-likelihood drops by about
  1). Moments of the draws cannot give this: when adapt runs in the neck the draws show the prior,
  not the likelihood.
- The local kernel's covariance is `epsilon^2 (L + Sigma^-1)^-1` at the current `Sigma`, in place
  of `epsilon^2 chains_var`. Epsilon is then a relative scale that no longer has to track the
  group level, and the 0.01 floor is not relied on.
- The global kernel's `N(chains_mu, chains_var)` component is replaced by the Gaussian
  approximation of the subject's conditional posterior at the current group level,
  `N((L + Sigma^-1)^-1 (l + Sigma^-1 mu), (L + Sigma^-1)^-1)` with `l` the linear term.
- The local kernel keeps at least a quarter of the sample-stage iterations.

`type = "single"` has no group level to follow and keeps `chains_var`.

**Tuning.** The local step's epsilon is adapted so that 3% of its particles beat the current value
(on the toy this is where ESS per iteration is maximal in 9, 25 and 35 dimensions; the old 25-30%
targets give steps 3-5x too small for the exact kernel); clamp [.01, 20] instead of the floors;
independence components are used at the scale of their estimated covariance (epsilon 1); burn's
`chains_var` epsilon, not its prior-variance one, is carried into adapt. In the exact kernels
epsilon is updated only in the iterations that used the local kernel, from that iteration's own
rate, with gain 0.1 and the update normalised by the target and clipped to [-1, 3]. (Updating it
every iteration from the rate accumulated over a window made it overshoot: over subjects it ended
anywhere between 0.01 and 5, which did not matter while it kept moving and does once it is
frozen.) preburn and burn keep the windowed counts, which restart every `step_size` iterations
(cumulative counts let epsilon run away). `update_epsilon_scale()` uses the square root of the
variance ratio (epsilon scales SDs). The `diag(.5)` fallback is replaced by the mean of the other
subjects' covariances, else the chain's previous one, else 0.1 x the prior variance diagonal.
`get_conditionals_single()` uses all (<= 750) adapt/sample draws rather than a random 250.

Everything above is switched by `options(emc.sampler = "legacy")`, which restores the old kernel in
every stage together with the old targets, floors, counts, fallback, 250-draw conditionals and the
sample-stage adaptation.

## 5. Reproducing it in EMC2

`vignettes/data/build_sampler_validity.R` simulates one subject (about 500 trials) from a
25-parameter LNR (`m, s ~ 0 + NWS:WF:lM, t0 ~ 1`, parameter values from a real participant), fits
it with both samplers using `fit()` defaults, and computes the Laplace SDs. Two minutes on three
cores; the results are cached.

```{r}
load("data/sampler-validity.RData")
tab <- t(sapply(summaries, function(s) c(
  `excess/k` = round(s$excess / s$k, 2),
  `SD/Laplace (median)` = round(median(s$pars$post_sd / s$pars$laplace_sd), 2),
  `SD/Laplace (min)` = round(min(s$pars$post_sd / s$pars$laplace_sd), 2),
  `SD/Laplace (max)` = round(max(s$pars$post_sd / s$pars$laplace_sd), 2),
  `max |z|` = round(max(abs(s$pars$z)), 1), `move rate` = round(s$move_rate, 2),
  `max Rhat` = round(s$max_rhat, 3), iterations = sum(s$iters))))
tab
```

```{r, fig.width = 7, fig.height = 4}
r <- sapply(summaries, function(s) s$pars$post_sd / s$pars$laplace_sd)
plot(r[, "legacy"], type = "h", lwd = 6, col = "grey60", ylim = c(0, max(r) * 1.05), xaxt = "n",
     xlab = "parameter", ylab = "posterior SD / Laplace SD", main = "25-parameter LNR, ~500 trials")
points(seq_len(nrow(r)) + .25, r[, "current"], type = "h", lwd = 6, col = "#0072B2")
abline(h = 1, lty = 2); axis(1, at = seq_len(nrow(r)), labels = FALSE)
legend("topleft", c("legacy sampler", "current sampler"), fill = c("grey60", "#0072B2"), bty = "n")
```

The legacy sampler reports Rhat ~1 and a move rate of 1.00 and is wrong by a third in every SD; the
current sampler's honest move rate is about .8. The per-stage tuning state (`epsilon`, `mix`,
`n_particles`) is in `attr(emc[[1]]$samples, "pm_settings")`:

```{r}
sapply(summaries, function(s) round(s$epsilon, 2))
```

With the legacy sampler the independence components end at an epsilon above 1 (1.2-1.6 from run
to run; wider than the posterior, which is where the over-dispersion comes from: the chain follows
them); with the current one they are fixed at 1 and only the local step is scaled (about
.4-.5 x `chains_var`).

To go further -- more parameters, more trials, `type = "standard"` -- change the design and
`n_trials` in the build script; the full grid, its runner and per-stage snapshots of the tuning
state are in `rating-work/sampler/grid/` (outside the package).

## 6. Evidence that the current sampler is calibrated

**Grid** (`rating-work/sampler/REPORT.md`, sections 1 and 5; 3 chains, `fit()` defaults, 1000 sample
iterations). Single subject, excess/k and median SD/Laplace, legacy -> current:

| k | N = 500 | N = 2000 | N = 8000 |
|---|---|---|---|
| 9 | 1.00 / 1.04 -> 0.90 / 0.99 | 1.09 / 1.06 -> 0.95 / 0.99 | 1.09 / 1.06 -> 0.99 / 1.00 |
| 25 | 1.68 / 1.34 -> 0.94 / 1.01 | 1.51 / 1.25 -> 0.97 / 0.99 | stuck -> 1.00 / 1.00 |
| 29 | 1.92 / 1.39 -> 0.92 / 0.96 | stuck -> 0.98 / 0.97 | stuck -> 0.99 / 1.00 |
| 35 | 1.92 / 1.38 -> 0.92 / 0.99 | stuck -> 0.92 / 0.92 | stuck -> 0.95 / 0.98 |

("stuck" = burn never converged, sample-stage move rate .01.) Rhat <= 1.10 in every current-sampler
cell, burn-in 100-300 iterations as before, and the 35-parameter x 8000-trial fit takes 22 minutes.
Eight subjects with `type = "standard"`: every cell with >= 2000 trials per subject is calibrated
(excess/k 0.91-1.02, SD ratio 0.89-0.99; legacy 1.27-1.64 and 1.09-1.23 at k >= 25). The
8-subject x 500-trial cells with 25-35 parameters do not reach Rhat < 1.1 under `fit()` defaults
with either sampler (see section 7).

**Simulation-based calibration.** The MTLNR SBC (8 single-subject cells x 500 replicates, one
15-subject cell x 100; `SBC-archive/2026-09-30_dev-sampler_MTLNR-sbc-regression`) rerun with the
current sampler: no KS p below .01, Fisher's combined p over the 68 parameter x cell tests .15 (the
legacy run: .051), 95% coverage .92-.98 centred on nominal (legacy ran high, .93-.98), |std_bias|
<= .17, hierarchical mu / Sigma ranks uniform; 3.5 h instead of 5.8 h.

**Real data.** The five Ratcliff-Starns models of `vignette("rating-models")`'s source paper
(24-35 parameters x 11 subjects, ~7600 trials each) all converge in the standard 1000 sample
iterations; EMC2's DIC ordering equals the paper's, EMC2 finds lower deviance minima than the
paper's DE-MCMC fits did, and the parameter estimates correlate >= .98 with them.

**After the 2026-10-01 changes of section 4** (fixed sample-stage kernel, proposals that follow
the group level; `rating-work/sampler/hier/stageH1b/REPORT.md`):

- *Conjugate model with a known posterior* (25 parameters x 20 subjects, 12 of them weakly
  identified, 20000 sample iterations x 6 chains from the truth and 6 from a collapsed start).
  The posterior of the weakly identified group SDs is reproduced from both starts: medians
  0.146-0.170 against 0.159 exact, share of draws in the funnel's neck 0.18-0.27 against 0.18-0.26,
  subject-level posterior SDs 0.97-1.01 x exact. With the group level held fixed the subject step
  alone is exact at the truth and in the collapsed state (SD ratio 1.00, about 2 iterations per
  effective draw; before the change it did not move in the collapsed state).
- *The grid's single-subject cells* (9, 25 and 35 parameters x 500 / 2000 / 8000 trials), rerun:
  excess/k 0.90-1.00, median SD/Laplace 0.98-1.03, Rhat <= 1.09. The 25-parameter cells take
  1.3-1.5x as long, because the number of particles is now set once, at the end of adapt.
- *A hierarchical cell with no funnel* (25 parameters x 20 subjects x 2000 trials): Rhat 1.02,
  excess/k 1.03, as before.

**Tests.** The suite passes (1434); the three snapshot tests in `test-fit.R` that pin specific
draws were re-recorded, as any change to the sampler requires.

## 7. The interweaving scale move (2026-10-02)

Section 6 found what remained once the particle step itself was repaired: the centred alternation
"subjects given the group level, group level given the subjects" mixes slowly in a funnel even
with a perfect subject step. On the conjugate model of `rating-work/sampler/hier/stageH1/
REPORT.md` one effective draw of a weak parameter's log group SD took about 500 iterations, and
leaving a collapsed state took about 400; `fit()`'s 1000 sample iterations were not enough. The
textbook remedy is interweaving / non-centred reparameterisation (Yu & Meng, 2011;
Papaspiliopoulos, Roberts & Sk&ouml;ld, 2007): after the centred group step, move in a
parameterisation where the group scale and the subjects' deviations are not conditionally
dependent.

`scale_move_standard()` (`R/variant_standard.R`; called from `run_stage()` in `R/sampling.R` after
the group step, in `adapt` and `sample`, `type = "standard"` only -- `do_scale <- pmwgs$type ==
"standard"`) does this per parameter `j`: a **scale move** proposes a log-scale change `delta`,
`alpha'_sj = mu_j + f (alpha_sj - mu_j)` for every subject (`f = exp(delta)`), `Sigma'_jj` scaled
by `f^2` and the Huang-Wand auxiliary variable `a_j` scaled by `1/f^2` (so the prior's own
hierarchy moves with it), accepted against the prior/Jacobian ratio `delta - (e^{2 delta} - 1) /
(A_j^2 a_j)` plus every subject's exact likelihood ratio; and, when there is no group-level design
(`do_loc <- is.null(sampler$group_designs) && length(pars$tmu) == p`), a **location move**
`mu'_j = mu_j + e`, `alpha'_sj = alpha_sj + e` for every subject, accepted the same way. Both run
on H1b's quadratic likelihood surrogate (`lik_prec`, the finite-difference Hessian already built
for the subject step) in `B` random parameter blocks of any size, each block's combined proposal
accepted or rejected against the subjects' *exact* likelihoods in one more `calc_ll_manager` call
per subject (delayed acceptance, so the move is exact whatever the surrogate's quality, not an
approximation) -- `B x n_subjects` extra single-row likelihood evaluations per iteration, about
0.1 ms each. The scale step size (Robbins-Monro, target .3 scaled as below) and `B` itself (kept where
the per-block acceptance is in [.35, .75]) adapt during `adapt`, are averaged over its tail the
same way the particle kernel's own tuning is (section 4's rule: nothing adapts once draws are
kept), and are frozen for `sample`. The acceptance the step sizes adapt on is the *realised* one:
a move counts only when the surrogate accepted it *and* its block then passed the exact check
(`acc_real` in `attr(samples, "scale_move")`; `acc_in` still counts the surrogate's). Adapting on
the surrogate's acceptance alone, as the first version did, tunes the step to the surrogate
whatever its quality: on forstmann's DDM `sv`/`SZ` (below), where the surrogate was wrong by
three orders of magnitude, its acceptance was a coin toss on the proposal's sign at any step, the
step ran away to 6-11 log-SD units and the exact check rejected nearly every block. With the
realised acceptance the exact check's rejections bound the step by what the exact likelihoods
accept; where the surrogate is exact (a Gaussian likelihood) every block passes and nothing
changes. Because a move shares the fate of its block, the Robbins-Monro target of the realised
acceptance is `.3 x r`, with `r` a running block acceptance (`r_block`: exponentially weighted,
weight .05 per sweep, starting at 1, frozen with the steps): a parameter whose moves are
innocuous then keeps a surrogate acceptance near .3, where a fixed target of .3 through blocks
that pass half the time pushed it to .6 and shrank every step to a quarter, which cost group-SD
mixing on the grid's hardest cell; a parameter whose moves make blocks fail more than the
average block still has its step shrunk. `r` enters the target no lower than .35, the block
acceptance below which the sweep is split into more blocks, so that when blocks cannot be split
further the target does not follow a collapsing block acceptance down (`test-scale-move.R`
checks this case with a surrogate that is flat in every parameter;
`rating-work/sampler/hier/stageH5b/REPORT.md` has the replicated comparison of the three
rules). `options(emc.scale_move =
FALSE)` turns the sweep off; the legacy sampler never runs it.

**Validation** (`rating-work/sampler/hier/stageH2/REPORT.md`, conjugate cell `p25`, 25 parameters
x 20 subjects). From both the truth and a collapsed start, group-SD quantiles (1-95%) equal the
exact reference to three decimals in every identifiability class, and the subject-level posterior
SD ratio is 0.997-1.003. The autocorrelation time of a weak parameter's log group SD falls from
about 500 (section 6) to 10-16, of its group mean to 4-5; median first escape from a collapsed
start falls from about 400 (exact) / 497 (the repaired particle step alone, section 6) to 28.5.
Under `fit()` defaults the reproducer (`k25_n20_N200`, which used to stop at max tries) converges
in 3 of 3 runs at the first check (1000-1200 sample iterations, alpha Rhat 1.05-1.09), at a
measured cost of 1.20x per sample iteration -- at the plan's own 20% budget line.

**Regression testing** (`rating-work/sampler/hier/stageH3/REPORT.md`) reran the package's grid,
an SBC cell and two real datasets under the sweep. It passes on simulated data with known truth:
11 of the grid's 12 `standard` cells and all 4 of section 6's funnelling cells meet Rhat < 1.1
(the twelfth, `m35_standard_500`, was already the grid's hardest cell pre-sweep and fails only
barely); the SBC cell it does not even touch (`diagonal-gamma`, where `do_scale` is `FALSE`) shows
no regression either. Two things did not survive contact with real data, and are open issues
(section 8), not resolved by this vignette:

- On forstmann (19-subject DDM), the scale move's step size ran away to 6-11 log-SD units on the
  weakly-identified `sv`/`SZ` parameters (every other parameter: 0.1-0.4) -- not a new failure
  (the pre-sweep sampler already struggled there) but a worse one, because a step size adapted on
  the surrogate's acceptance has nothing to stop it growing against a flat likelihood direction.
  The runaway is fixed by adapting on the realised acceptance (above), but forstmann is not
  (`rating-work/sampler/hier/stageH5/REPORT.md`): the step now shrinks instead, because the
  surrogate for `sv`/`SZ` is wrong by three orders of magnitude. `lik_precision()`'s step search
  oscillates between a step on the likelihood's flat stretch and one that straddles the DDM's
  floor at its bounds (`sv`, `SZ` < .01), ending on either side, and reads the floor as a
  curvature of 10^3-10^4 with a gradient to match. The same surrogate is the subject step's local
  proposal, which is why `sv` already failed to converge there before the sweep existed. The
  search now brackets its step and bisects, and gives a parameter whose step never settles a flat
  surrogate (`test-lik-precision.R`); on forstmann `SZ` converges (1.03 at 5000 draws) and `sv`
  goes from 5.3 to 1.2-1.7 at `fit()`'s limit and 1.18 at 5000 draws with every chain crossing the
  whole of its posterior, every other parameter below 1.01, in half the time. What is left on `sv`
  is the random walk of its group mean across a two-log-unit plateau where the likelihood is
  indifferent to `sv` (ending at the model's floor at `sv` = .01): about 200 iterations per
  traverse, so Rhat < 1.1 needs of the order of 10^4 draws. That is the model and prior, not the
  sampler. Where the likelihood is quadratic (the conjugate model) the change is a no-op; on the
  `k25` funnel cells the surrogate is entry-for-entry identical.
- On the Eisenberg Simon data (518 subjects, no funnel present -- mu/sigma Rhat both fine), the
  sweep makes subject-level (alpha) Rhat *worse* than the pre-sweep sampler (1.212 vs 1.098) at
  4x the cost, with the block-acceptance diagnostic below its target window -- the quadratic
  surrogate the delayed-acceptance step relies on may fit this model/N poorly.
- The sweep's own ~20% cost estimate, measured on the `k25` reproducer, does not generalise: every
  one of the grid's 12 `standard` cells exceeds it (1.56x-11.57x per sample iteration), including
  9-parameter cells that never funnel. It stays close to ~20% only on cells that actually funnel.

**Many subjects (2026-10-05; `rating-work/sampler/hier/stageH6/REPORT.md`, `stageH6b/REPORT.md`,
`stageH6b/REPORT_partB.md`).** The Eisenberg Simon regression above (RDM, 518 subjects, 8 parameters)
turned out not to be the sweep: with the sweep off it was the same (20 of 20 fits above 1.1 at
`fit()`'s maximum tries, minimum subject-level ESS 14 per 1000 draws). The cause was section 4's fixed
sample-stage kernel, built when `adapt` stopped on `min_unique` after about 100 iterations, from draws
in which a few subjects' chains still sat apart; those subjects' kernels were poor for the rest of the
fit and set its Rhat. The mean Rhat over a window (burn's criterion) does not show this; the largest
does. Four changes, replicated five times per cell on Gadi:

- *Converged adapt.* `adapt` also waits until the largest Rhat of alpha over the last 250 iterations
  (the window the kernel is built from) is below 1.2, from 250 iterations on; it gives up after three
  checks without a new minimum or at 1000 (`adapt_converged()`, `adapt_stop()` in `R/fitting.R`;
  `options(emc.adapt_converge = FALSE)` switches it off). The efficient proposals are built from
  that window.
- *Likelihood precision from the draws* (`lik_precision_draws()`). The particle step's conditional
  proposal used the finite-difference curvature at one point, two to three times too narrow for the
  skewed posteriors of the worst subjects (`t0`, `B`). It now uses the window's draws: their inverse
  covariance minus the window's mean group precision, in the directions where that is clearly
  positive, and the finite differences (capped at what the draws allow) elsewhere. The subtraction
  is only right if the group precision is about constant over the window, so this is used only
  where the largest coefficient of variation of the group precision over the window is below .5
  (`lik_prec_draws_cv`): .17-.33 on Eisenberg and on a simulated 300-subject cell, 8-20 for the worst
  parameter of the ladder's funnel cells, where an earlier version that used it everywhere halved
  the conditional proposal's acceptance and cost a third of the subject-level ESS.
- *A gate on the sweep.* With hundreds of subjects the surrogate's error has the same sign in nearly
  every subject and adds up, so moves of well-identified parameters pass the exact check only at
  steps far below the posterior SD of what they move, while each block still costs one likelihood
  per subject. A move is dropped for the rest of the fit when the sweep is already split into one
  block per parameter and its expected squared jump per sweep is under a tenth of the group-level
  Gibbs step's own (`scale_move_gate()`, decided from 300 adapt sweeps on). On Eisenberg it keeps
  four scale and three location moves; on the ladder cells it never fires. The steps start at each
  move's conditional scale on the surrogate.
- *The exact check in-process.* The block's exact check forked one process per core; for cheap
  likelihoods forking cost more than the likelihoods (0.26-0.34 s against 0.10-0.13 s per block on
  Eisenberg). The first checks are timed both ways and the faster is kept (`check_serial` in
  `attr(samples, "scale_move")`).

Results with `fit()` defaults (five runs each): Eisenberg subject-level Rhat 1.10 [1.08, 1.16] (3 of 5
below 1.1), ESS 5% quantile / minimum 211 / 48 per 1000 draws, 63 minutes (the ungated sweep took 150);
against the Stan reference of `rating-work/sampler/stan/REPORT_refs.md` no group-level quantity is
more than 4 Monte-Carlo SEs off (two between 3 and 4, on `B` and `t0`). The `k25` funnel cells and
`m35_standard_500` mix as well as or better than before (conditional-proposal acceptance .29-.33
against .14-.21, a third to a half as many particles) and run 2-4 times faster; `p25` reproduces the
exact posterior. forstmann (DDM) is 1.3 times slower: its `sv` plateau keeps adapt's window Rhat at
1.2-2.2, so adapt runs 600-1000 iterations.

The earlier sampler that passed Eisenberg (`002a5c6f`, which kept re-estimating its proposals during
sampling, section 4) did so with a biased posterior: against Stan it was off on 10 of 16 group means
and log SDs (up to 7.5 Monte-Carlo SEs) and too narrow (SD ratios 0.84-1.00), and the subjects with
heavy posterior shoulders (below) were under-sampled.

## 8. What is still open

An earlier version of this section said that the hierarchical fits which do not converge
(few trials per weakly identified parameter: 8-40 subjects x 200-500 trials x 25-35 parameters,
alpha Rhat 1.4-1.9 under `fit()` defaults) were stuck in a funnel because of the group-level Gibbs
step, not the particle kernel. That was wrong. On a conjugate model with a known posterior
(`rating-work/sampler/hier/stageH1/REPORT.md`) EMC2's group step with an exact draw of the subject
parameters reproduces the exact posterior, and the failure was the particle step: the sample
stage's continued adaptation made chains sink into the funnel's neck from healthy states, and
the local proposal froze once they were there (both repaired, section 4); what was left after
that repair was the centred alternation's own slow mixing in the funnel, addressed by the
interweaving sweep of section 7.

What remains, as of this writing:

- **Eisenberg Simon: what is left after section 7's many-subject changes.** Two runs in five still
  stop just above 1.1. A few subjects (`s095` `B`, `s273` `t0`, `s084` `v_lMd`) have 3-5% of their
  posterior mass in a shoulder several SDs from the mode (Stan: skewness 2.5-3.6; not the
  likelihood floor): a chain with Gaussian proposals enters it rarely and stays a few hundred
  iterations, so at 1000-2000 draws a run either misses the shoulder or over-weights it. And the
  group means of `v` and `v_lMd` sit on a ridge (the matching accumulator's rate is identified,
  their difference barely; autocorrelation time 45-100 under every build), which single-parameter
  moves cannot follow. Both need new moves (a local kernel with a mixture of scales for the
  shoulders; a joint group-level move for the ridge). forstmann's `sv` stays slow for the same kind
  of reason (its group mean on a likelihood plateau), which no step-size change addresses. One
  Eisenberg fit in 54 never left the likelihood floor after `preburn` (one chain with every
  subject's `v_iMd` at its bound): open.
- **The ~20% cost estimate is per-model-family, not general** (section 7): expect more overhead
  than that on models that are not shaped like the `k25` reproducer, even where there is no
  funnel.
- **The location move only runs without a group-level design** (`do_loc`, section 7); with one,
  only the scale move runs, so a funnel in the group means under a group design is not yet
  addressed.
- **The scale move is diagonal only** (`Sigma'_jj` scaled independently per parameter), and a
  correlated collapse now has a reproducer: a simulated 300-subject RDM cell with a 2 x 2 design
  (`stageH6b/many_cells.R`, `variant = "C"`), in which only `v_Cd + v_lMd:Cd / 2` is identified per
  subject and the group SDs of the two parameters are not. Every build fails it (subject-level Rhat
  1.5-2.6), the fixed-kernel builds worse than `002a5c6f`. Scaling one parameter's deviations moves
  the identified combination, so this needs a joint (rotated) group-level move; not implemented.
- **Only `type = "standard"` runs the sweep** (`do_scale`); `diagonal` and `blocked` share enough
  of the code path that they probably only need the call site enabled, but this is unconfirmed.
  `factor`, `infnt_factor` and `SEM` are untested for the move; `diagonal-gamma` was incidentally
  regression-tested clean on everything *except* the move itself, which never runs for it.
- **No sweep in burn**; `init()` still starts every chain in the neck (subjects bunched at the
  group mean, group SD about 0.14 for every parameter), and `preburn`/`burn` keep the old search
  step.
- **The exact reference is R-only for the conjugate model** (`stageH1/exact_gibbs.R`); Stan (NUTS)
  references now exist for it and for the `k25` cells, forstmann and Eisenberg
  (`rating-work/sampler/stan/REPORT_refs.md`). The systematic comparison of the final build with
  all of them is still to come.
- **Dispersed starts** (as an alternative or supplement to fixing the mixing) have not been tried.

## References

Neal, R. M. (2011). MCMC using ensembles of states for problems with fast and slow variables such
as Gaussian process regression. arXiv:1101.0387.

Papaspiliopoulos, O., Roberts, G. O., & Sk&ouml;ld, M. (2007). A general framework for the
parametrization of hierarchical models. Statistical Science, 22(1), 59-73.

Tjelmeland, H. (2004). Using all Metropolis-Hastings proposals to estimate mean values. Statistics
No. 4/2004, Norwegian University of Science and Technology.

Yu, Y., & Meng, X.-L. (2011). To center or not to center: That is not the question -- an
ancillarity-sufficiency interweaving strategy (ASIS) for boosting MCMC efficiency. Journal of
Computational and Graphical Statistics, 20(3), 531-570.
