---
title: "Latent Stop and Go Finishing Times"
description: "Simulating SSRT and go-race finishing times from stop-signal models with make_stop_data()"
author: "Andrew Heathcote"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{"Latent Stop and Go Finishing Times"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
rm(list = ls())
library(EMC2)
set.seed(1)
knitr::opts_chunk$set(fig.width = 7, fig.height = 4, message = FALSE)
```

## Introduction

In the race models of the stop-signal task, `SSEXG` and `SSRDEX`, a stop racer
and one or more go racers run on each stop trial, and the observed response
depends on which finishes first. The quantity of main interest, the stop-signal
reaction time (SSRT), is never observed: on a successful stop we only know the
stop racer finished before the go racers, and on a signal-respond trial we only
know it did not.

`make_stop_data()` simulates the race and returns, for every simulated trial,
the finishing times that produced it:

* `SSRT`: the stop racer's finishing time, measured from stop-signal onset;
* `goRT` and `goR`: the finishing time (from go onset) and identity of the
  winner of the go race.

It can simulate from a parameter vector, like `make_data()`, or from the
posterior of a fitted model, like `predict()`. The posterior version gives
posterior distributions of SSRT and of the go-race winner's time under the
fitted model, including all the features that model has: lower bounds on the
finishing-time distributions, trigger and go failures, stop-triggered
accumulators, trends, and censoring and truncation of the data.

This vignette describes what `make_stop_data()` returns, how it works, how to
use it, and how it has been tested.

## What it returns

The output is the simulated data set, the same as `make_data()` or `predict()`
gives (with a `postn` column for posterior draws), plus these columns:

| Column | Content |
|--------|---------|
| `goR` | Winning go accumulator (a factor with the response levels). `NA` on a go failure. |
| `goRT` | Finishing time of the go-race winner, from go onset. `Inf` on a go failure. |
| `SSRT` | Stop racer's finishing time, from stop-signal onset. `Inf` on a trigger failure; `NA` on go trials (no stop signal) unless `stop_on_go_trials = TRUE`. |
| `goMissingness` | Censoring of `goRT`, coded like `missingness`: `NA` observed, `1` below `LC`, `2` above `UC`. |

`goRT` and `goR` are reported on every trial, including stop trials where the
stop racer won and so no go response was made. That is the point of the
function: it gives the go time that the stop process inhibited.

## How it works

`make_stop_data()` does not have its own simulator. It uses the stop-signal
models' random functions, which `make_data()` and `predict()` also use. For each trial,
those functions draw a finishing time for every racer into one matrix:
the stop racer, the go racers and any stop-triggered racers. Stop
finishing times are drawn relative to stop-signal onset. The stop-signal delay
(SSD) is then added to the stop racer, and the response is whichever racer
finishes first.

With `make_stop_data()` the same functions also return the stop and go times
from that matrix, before SSD is added. Several properties follow from this:

* **The latent times are the ones that produced the data.** On a go trial
  with a response, `rt == goRT`. On a signal-respond trial, `rt == goRT` and
  `goRT < SSD + SSRT`. On a stop trial with no go response, either the go
  process failed (`goRT = Inf`) or `goRT > SSD + SSRT`.
* **Lower bounds are respected.** `SSRT` is drawn from the ex-Gaussian
  distribution with lower bound `exgS_lb`. The bound applies to the time from stop-signal
  onset, as in the likelihood, so `SSRT > exgS_lb` while `SSD + SSRT` can be
  smaller. For `SSEXG`, go finishing times have lower bound `exg_lb`. These
  bounds are parameters of the finishing-time distributions, distinct from
  truncation of the data (below).
* **Failures are explicit.** A trigger failure (probability `tf`) gives
  `SSRT = Inf`; a go failure (probability `gf`) gives `goRT = Inf` and
  `goR = NA`.
* **Stop-triggered accumulators** (factor `lI`) are not part of the go race:
  `goR` never names one.
* **Trends** on go and/or stop parameters are applied trial by trial, because
  the latent times are drawn from each trial's own parameters. When a trend
  covariate depends on behaviour, or the SSD staircase is re-run, the data
  are simulated trial by trial, as in `predict()`.
* **Truncation and censoring follow the data.** Censoring and truncation
  settings (`TC` in `design()` or `make_data()`, or `LT`/`LC`/`UC`/`UT` columns
  in the data; see `?make_missing`) are applied as for the observed data.
  A truncated trial is removed with its latent times. The go race is censored
  at the same `LC`/`UC` bounds as `rt`: a censored `goRT` is `NA`, and
  `goMissingness` records the direction. A go failure counts as a response
  later than a finite `UC`. Set `censor_go = FALSE` for the uncensored go
  times.
* **The simulated data do not change.** Reading the latent times takes no
  extra random numbers, so for a given seed `make_stop_data()` returns
  exactly the `R`, `rt` and `SSD` that `make_data()` does. With
  `stop_on_go_trials = TRUE`, the extra stop draws on go trials are made
  last, so they do not change the data either.

## Using it with a parameter vector

We use an ex-Gaussian race with two go racers, one matching the stimulus and
one not. `make_ssd()` adjusts SSD with a staircase.

```{r}
design_ss <- design(model = SSEXG,
  factors = list(subjects = 1, S = c("left", "right")),
  Rlevels = c("left", "right"),
  matchfun = function(d) as.numeric(d$S) == as.numeric(d$lR),
  formula = list(mu ~ lM, sigma ~ 1, tau ~ 1, muS ~ 1, sigmaS ~ 1, tauS ~ 1,
                 gf ~ 1, tf ~ 1), report_p_vector = FALSE)
p_true <- c(mu = log(.6), mu_lMTRUE = log(.8), sigma = log(.05), tau = log(.2),
            muS = log(.2), sigmaS = log(.03), tauS = log(.05),
            gf = qnorm(.05), tf = qnorm(.1))
sim <- make_stop_data(p_true, design_ss, n_trials = 2000,
                      functions = list(SSD = make_ssd()))
head(sim[, c("S", "SSD", "R", "rt", "goR", "goRT", "SSRT")])
```

The lower bounds `exg_lb` and `exgS_lb` take their default value of 0.05.

### The stop finishing-time distribution

On stop trials where the stop racer was triggered, `SSRT` follows the
ex-Gaussian with lower bound `exgS_lb`. The curve is that density, computed
directly from the parameters.

```{r}
dexg <- function(x, mu, sigma, tau)
  exp(sigma^2 / (2 * tau^2) - (x - mu) / tau) / tau * pnorm((x - mu) / sigma - sigma / tau)
pexg <- function(q, mu, sigma, tau)
  pnorm((q - mu) / sigma) -
    exp(sigma^2 / (2 * tau^2) - (q - mu) / tau) * pnorm((q - mu) / sigma - sigma / tau)
dtexg <- function(x, mu, sigma, tau, lb)
  ifelse(x > lb, dexg(x, mu, sigma, tau) / (1 - pexg(lb, mu, sigma, tau)), 0)

stop_trial <- is.finite(sim$SSD)
ssrt <- sim$SSRT[stop_trial & is.finite(sim$SSRT)]
hist(ssrt, breaks = 50, freq = FALSE, main = "SSRT", xlab = "Seconds from stop-signal onset")
curve(dtexg(x, .2, .03, .05, .05), add = TRUE, lwd = 2, col = "red")
c(trigger_failure_rate = mean(is.infinite(sim$SSRT[stop_trial])), min_SSRT = min(ssrt))
```

### The go race

`goRT` is the minimum of the go racers' finishing times. On a stimulus
`left` trial the matching racer has `mu = .6 * .8` and the other `mu = .6`,
so the density of the winner's time is $f_1 S_2 + f_2 S_1$, where $S = 1 - F$.

```{r}
Sexg <- function(x, mu) {
  out <- 1 - (pexg(x, mu, .05, .2) - pexg(.05, mu, .05, .2)) / (1 - pexg(.05, mu, .05, .2))
  ifelse(x > .05, out, 1)
}
dwin <- function(x) dtexg(x, .48, .05, .2, .05) * Sexg(x, .6) + dtexg(x, .6, .05, .2, .05) * Sexg(x, .48)
go_left <- sim$goRT[sim$S == "left" & is.finite(sim$goRT)]
hist(go_left, breaks = 50, freq = FALSE, main = "Go-race winner (stimulus left)",
     xlab = "Seconds from go onset")
curve(dwin(x), add = TRUE, lwd = 2, col = "red")
```

### Seeing the race

Each stop trial is a point: the go winner's time against the stop racer's
finishing time on the go clock (`SSD + SSRT`). Points below the diagonal are
signal-respond trials, where the go racer finished first and a response was made. Points
above it are successful stops.

```{r}
st <- stop_trial & is.finite(sim$goRT) & is.finite(sim$SSRT)
responded <- !is.na(sim$R[st])
plot(sim$SSD[st] + sim$SSRT[st], sim$goRT[st], pch = 16, cex = .5,
     col = ifelse(responded, "black", "grey60"),
     xlab = "SSD + SSRT", ylab = "goRT")
abline(0, 1, col = "red")
legend("topleft", c("signal-respond", "successful stop"),
       pch = 16, col = c("black", "grey60"), bty = "n")
```

### SSRT on go trials

No stop signal is presented on go trials, so `SSRT` is `NA` there. Set
`stop_on_go_trials = TRUE` to draw a stop racer on every trial. This gives more
SSRT draws for estimating its distribution, and the simulated
data are unchanged.

```{r}
set.seed(2)
a <- make_data(p_true, design_ss, n_trials = 200, functions = list(SSD = make_ssd()))
set.seed(2)
b <- make_stop_data(p_true, design_ss, n_trials = 200,
                    functions = list(SSD = make_ssd()), stop_on_go_trials = TRUE)
identical(a[, c("R", "rt", "SSD")], b[, c("R", "rt", "SSD")])
sum(is.na(b$SSRT))
```

## Censoring and truncation

Suppose responses faster than 0.4 s and slower than a 0.7 s deadline are
censored. `rt` is then censored, and so is the go-race winner, at the same
bounds.

```{r}
set.seed(3)
cen <- make_stop_data(p_true, design_ss, n_trials = 500,
                      functions = list(SSD = make_ssd()), TC = list(LC = .4, UC = .7))
table(goMissingness = cen$goMissingness, useNA = "ifany")
range(cen$goRT, na.rm = TRUE)
```

On go trials `goMissingness` equals `missingness`. On stop trials the go race
is censored even when the stop racer won: whatever the stop process did, the
go response would not have been observed in the window. Use `censor_go = FALSE`
to keep the uncensored go times.

Truncation (`LT`, `UT`) removes trials whose observed response falls outside
the window. Their latent times are removed with them, so the latent times are
for the same trials as the observed data.

## Trends

Latent times are drawn from each trial's own parameters, so trends carry
through. Here the mean of the stop racer's Gaussian component rises with SSD:
`muS = 0.2 + 0.2 * min(1, 5 * SSD)`.

```{r}
trend <- make_trend(make_base("muS", "lin", make_kernel("SSD", "slin_incr"),
                              phase = "posttransform"))
design_tr <- design(model = SSEXG,
  factors = list(subjects = 1, S = c("left", "right")),
  Rlevels = c("left", "right"),
  matchfun = function(d) as.numeric(d$S) == as.numeric(d$lR),
  trend = trend, transform = list(func = c(muS.w = "exp")),
  formula = list(mu ~ lM, sigma ~ 1, tau ~ 1, muS ~ 1, sigmaS ~ 1, tauS ~ 1,
                 gf ~ 1, tf ~ 1), report_p_vector = FALSE)
p_tr <- sampled_pars(design_tr, doMap = FALSE)
p_tr[] <- c(p_true, muS.k_sat = log(5), muS.w = log(.2))[names(p_tr)]
ssd_grid <- function(d) rep(c(Inf, .05, Inf, .1, Inf, .15, Inf, .3), length.out = nrow(d))
sim_tr <- make_stop_data(p_tr, design_tr, n_trials = 2000, functions = list(SSD = ssd_grid))
ok <- is.finite(sim_tr$SSD) & is.finite(sim_tr$SSRT)
round(tapply(sim_tr$SSRT[ok], sim_tr$SSD[ok], mean), 3)
```

The mean SSRT rises with SSD up to SSD = 0.2, where the trend reaches its
maximum.

## Using it with a fitted model

For a fitted model, `make_stop_data()` takes `n_post` draws from the posterior
and simulates the latent times for each. It has the same arguments as `predict()`.
By default the observed trials and SSDs are used. With
`conditional_on_data = FALSE`, a staircase given to `design()` is re-run
trial by trial.

We fit `SSEXG` to 500 trials simulated from `p_true` (this takes a few
minutes, so the fitted object is loaded below).

```{r, eval = FALSE}
dat <- make_data(p_true, design_ss, n_trials = 250, functions = list(SSD = make_ssd()))
emc <- make_emc(dat, design_ss, type = "single")
emc <- fit(emc, fileName = "data/make_stop_data.RData")
```

```{r}
load("data/make_stop_data.RData")
post <- make_stop_data(emc, n_post = 100, stop_on_go_trials = TRUE)
dim(post)
```

`stop_on_go_trials = TRUE` gives an SSRT on every trial (about 500 per
posterior draw rather than the 125 or so stop trials), which smooths the
per-draw distributions below. Statistics about the race itself use the stop trials only.

### Posterior distribution of SSRT

Each posterior draw gives an SSRT distribution. Below, the grey curves are
densities for 20 draws and the red curve is the true distribution.

```{r}
ok_ssrt <- is.finite(post$SSRT)
plot(NULL, xlim = c(0, .6), ylim = c(0, 10), xlab = "SSRT", ylab = "Density",
     main = "Posterior SSRT distributions")
for (i in 1:20) {
  x <- post$SSRT[ok_ssrt & post$postn == i]
  lines(density(x, from = .05), col = "grey70")
}
curve(dtexg(x, .2, .03, .05, .05), add = TRUE, lwd = 2, col = "red")
```

With about 125 stop trials the shape of the SSRT distribution is poorly
constrained by the data. `muS`, `sigmaS` and `tauS` trade off against each
other, so the posterior distributions are much wider than the true one. Its
location is well constrained. For example, the posterior of mean SSRT (which,
because of the lower bound, is not exactly `muS + tauS`) covers the true
value:

```{r}
mean_ssrt <- tapply(post$SSRT[ok_ssrt], post$postn[ok_ssrt], mean)
true_mean <- integrate(function(x) x * dtexg(x, .2, .03, .05, .05), .05, 2)$value
round(c(true = true_mean, quantile(mean_ssrt, c(.025, .5, .975))), 3)
```

The same can be done for the go race. For example, the proportion of stop
trials on which the stop process inhibited a go response that would otherwise
have been made (the go racer finished, but after the stop racer), and the mean
time of those inhibited go responses:

```{r}
stop_post <- is.finite(post$SSD)
inhibited <- stop_post & is.finite(post$goRT) & is.na(post$R)
p_inhibited <- tapply(inhibited[stop_post], post$postn[stop_post], mean)
rt_inhibited <- tapply(post$goRT[inhibited], post$postn[inhibited], mean)
round(rbind(p_inhibited = quantile(p_inhibited, c(.025, .5, .975)),
            mean_goRT_inhibited = quantile(rt_inhibited, c(.025, .5, .975))), 3)
```

## How it was tested

The tests are in `tests/testthat/test-make_stop_data.R`. They check:

1. **Distributions.** For `SSEXG` and `SSRDEX`, Kolmogorov-Smirnov tests compare
   `SSRT` with the ex-Gaussian distribution with lower bound `exgS_lb`, and `goRT` with the
   distribution of the minimum of the go racers (ex-Gaussian with lower bound
   `exg_lb` for `SSEXG`; `t0` plus Wald for `SSRDEX`). The comparison distributions
   are the ones the likelihood uses. The bounds are set high
   enough to matter (`exgS_lb = 0.15`), and every `SSRT` exceeds it.
2. **Exact consistency with the simulated data.** On every trial, for both
   models: `rt == goRT` and `R == goR` for go responses; `goRT < SSD + SSRT`
   on signal-respond trials; `goRT > SSD + SSRT` or `goRT = Inf` when no go
   response was made on a stop trial; `SSRT` is `NA` on go trials.
3. **Failures.** The proportions of `SSRT = Inf` on stop trials and of
   `goRT = Inf` match `tf` and `gf`.
4. **Stop-triggered accumulators.** With a stop-triggered response option,
   such responses occur but are never the go winner, and check 2 still holds.
5. **No change to the data.** With the same seed, `make_data()` and
   `make_stop_data()` (including `stop_on_go_trials = TRUE`) return identical
   `R`, `rt`, `SSD` and `missingness`, with staircase SSDs.
6. **Censoring and truncation.** `goRT` below `LC` or above `UC` is `NA` with
   `goMissingness` 1 or 2 and `goR` is `NA`; uncensored values equal the
   `censor_go = FALSE` values; on go trials `goMissingness` equals
   `missingness`; go failures stay `Inf` without a finite `UC`; with `LT`/`UT`
   the same trials are removed as by `make_data()`.
7. **Trends.** With a trend on `muS` over SSD, `SSRT` follows the trended
   distribution (Kolmogorov-Smirnov test). With a covariate trend on the go `mu`,
   `goRT` increases with the covariate. Check 2 holds in both cases.
8. **Fitted models.** On an `emc` object, both the default path (observed
   SSDs, which are reproduced) and the trial-by-trial path
   (`conditional_on_data = FALSE`, staircase re-run) return the latent columns
   and satisfy check 2 for every posterior draw.
9. **Model check.** Models other than `SSEXG` and `SSRDEX` are rejected.

The existing stop-signal, censoring and trend test files, and the full package
test suite, pass with these changes.
